---
title: "Transition Probability Heatmap"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load Libraries

```{r libraries}
library(ggplot2)
library(reshape2) # for melt()
```

```{r path}
path <- "/w5home/bmoore/Pierre_sc_zebrafish/output_seurat2ann_20260224_121120/realtime_moscot_recovered_20260325_093911/"
setwd(path)

# ── SET TIME POINTS HERE ──────────────────────────
# Change these two values to switch between transitions
time_from <- 3
time_to <- 4
```

## Read Data

Update the `csv_path` variable to point to your CSV file.

```{r read-data}
csv_file <- paste0("transitions_", time_from, "_", time_to, "_res0.75.csv")

# Read CSV — first column is row names
mat <- read.csv(csv_file, row.names = 1, check.names = FALSE)

# Convert to matrix then melt to long format for ggplot
mat <- as.matrix(mat)
df_long <- melt(mat)
colnames(df_long) <- c("Source", "Target", "Probability")

# Ensure cluster ordering is preserved (numeric order)
df_long$Source <- factor(df_long$Source, levels = rev(rownames(mat)))
df_long$Target <- factor(df_long$Target, levels = colnames(mat))
```

## Plot Heatmap

```{r heatmap, fig.width=10, fig.height=9}
p1 <- ggplot(df_long, aes(x = Target, y = Source, fill = Probability)) +
    geom_tile(color = "grey30", linewidth = 0.15) +
    scale_fill_viridis_c(
        option = "inferno",
        name   = "Transition\nProbability",
        limits = c(0, 1)
    ) +
    labs(
        title = "Cell-State Transition Probabilities",
        subtitle = paste0("Time ", time_from, " → Time ", time_to, "  |  Resolution 0.75"),
        x = paste0("Target Cluster (Time ", time_to, ")"),
        y = paste0("Source Cluster (Time ", time_from, ")")
    ) +
    theme_minimal(base_size = 13) +
    theme(
        plot.title       = element_text(face = "bold", hjust = 0.5),
        plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
        axis.text.x      = element_text(angle = 45, hjust = 1),
        panel.grid       = element_blank(),
        legend.position  = "right"
    ) +
    coord_fixed() # square tiles
```
```{r make_pdf}
ggsave(p1, filename = paste0("heatmap1_transitions_", time_from, "_", time_to, ".pdf"), width = 10, height = 9, dpi = 600)
```

### Optional: Log-scaled version

Very small transition probabilities are hard to see on a linear scale. A log10 transform can help reveal structure in the off-diagonal entries.

```{r heatmap-log, fig.width=10, fig.height=9}
# Replace exact zeros with NA so log transform doesn't produce -Inf
df_log <- df_long
df_log$Probability[df_log$Probability == 0] <- NA

p2 <- ggplot(df_log, aes(x = Target, y = Source, fill = Probability)) +
    geom_tile(color = "grey30", linewidth = 0.15) +
    scale_fill_viridis_c(
        option    = "inferno",
        name      = "Transition\nProbability",
        trans     = "log10",
        na.value  = "black",
        labels    = scales::label_scientific()
    ) +
    labs(
        title = "Cell-State Transition Probabilities (log\u2081\u2080 scale)",
        subtitle = paste0("Time ", time_from, " → Time ", time_to, "  |  Resolution 0.75"),
        x = paste0("Target Cluster (Time ", time_to, ")"),
        y = paste0("Source Cluster (Time ", time_from, ")")
    ) +
    theme_minimal(base_size = 13) +
    theme(
        plot.title       = element_text(face = "bold", hjust = 0.5),
        plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
        axis.text.x      = element_text(angle = 45, hjust = 1),
        panel.grid       = element_blank(),
        legend.position  = "right"
    ) +
    coord_fixed()
```

```{r make_pdf}
ggsave(p2, filename = paste0("heatmap2_transitions_", time_from, "_", time_to, ".pdf"), width = 10, height = 9, dpi = 600)
```

### Off-diagonal heatmap (diagonal masked)

Set the diagonal (cluster *i* → cluster *i*) to `NA` so only cross-cluster transitions are visible. The color scale auto-ranges to the off-diagonal values for better contrast.

```{r heatmap-offdiag, fig.width=10, fig.height=9}
# Copy the matrix and blank out the diagonal
mat_offdiag <- mat
diag(mat_offdiag) <- NA

# Melt to long format
df_offdiag <- melt(mat_offdiag)
colnames(df_offdiag) <- c("Source", "Target", "Probability")

# Preserve cluster ordering
df_offdiag$Source <- factor(df_offdiag$Source, levels = rev(rownames(mat)))
df_offdiag$Target <- factor(df_offdiag$Target, levels = colnames(mat))

p3 <- ggplot(df_offdiag, aes(x = Target, y = Source, fill = Probability)) +
    geom_tile(color = "grey30", linewidth = 0.15) +
    scale_fill_viridis_c(
        option   = "inferno",
        name     = "Transition\nProbability",
        na.value = "grey90"
    ) +
    labs(
        title = "Off-Diagonal Transition Probabilities",
        subtitle = paste0("Time ", time_from, " → Time ", time_to, "  |  Resolution 0.75  |  Diagonal masked"),
        x = paste0("Target Cluster (Time ", time_to, ")"),
        y = paste0("Source Cluster (Time ", time_from, ")")
    ) +
    theme_minimal(base_size = 13) +
    theme(
        plot.title       = element_text(face = "bold", hjust = 0.5),
        plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
        axis.text.x      = element_text(angle = 45, hjust = 1),
        panel.grid       = element_blank(),
        legend.position  = "right"
    ) +
    coord_fixed()

p3
```

```{r save_offdiag_pdf}
ggsave(p3, filename = paste0("heatmap3_offdiag_transitions_", time_from, "_", time_to, ".pdf"), width = 10, height = 9, dpi = 600)
```
