---
title: "Uniquely Expressed Genes Per Cluster"
author: "Antigravity Coding Assistant"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(cowplot)
```

# Introduction
This notebook loads a Seurat object, calculates the number of uniquely expressed genes in each cluster, and generates a scatter plot showing the cluster on the X-axis and the number of uniquely expressed genes on the Y-axis.

We implement two different definitions of "uniquely expressed genes":
1. **Method A (Marker-Based):** Genes that are identified as significant positive markers for a single cluster using `FindAllMarkers()`, and do not appear as markers for any other cluster.
2. **Method B (Expression-Fraction-Based):** Genes that are expressed in more than a minimum percentage of cells ($P_{min}$) in a given cluster, regardless of their expression in other clusters.

# Configuration & Load Data
Specify the path to the Seurat object and any parameter thresholds.

```{r config}
# File paths
SEURAT_PATH <- "/w5home/bmoore/Pierre_sc_zebrafish/output_subset-recluster_20260204_094842/seuratobj_all_reclustered.rds"
OUTPUT_DIR <- file.path("/w5home/bmoore/Pierre_sc_zebrafish/output_subset-recluster_20260204_094842",
                        paste0("output_unique_genes_", format(Sys.time(), "%Y%m%d_%H%M%S")))
dir.create(OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)

# Method A (Marker) Thresholds
LOGFC_THRESH <- 0.25
MIN_PCT <- 0.10
P_VAL_ADJ_THRESH <- 0.05

# Method B (Expression-Fraction) Thresholds
P_MIN <- 0.05 # Minimum expression fraction in target cluster (other-cluster expression not restricted)
```

```{r load_data}
cat("Loading Seurat object from:", SEURAT_PATH, "\n")
seurat_obj <- readRDS(SEURAT_PATH)
print(seurat_obj)

# Set the active identity to the desired cluster resolution
if ("seurat_clusters_res0.75" %in% colnames(seurat_obj@meta.data)) {
  Idents(seurat_obj) <- "seurat_clusters_res0.75"
  cat("Set active identity to seurat_clusters_res0.75\n")
} else {
  warning("seurat_clusters_res0.75 not found in metadata, using default active identity.")
}

cat("Active identities:\n")
print(table(Idents(seurat_obj)))
```

---

# Method A: Marker-Based Uniquely Expressed Genes
Using Seurat's `FindAllMarkers`, we identify statistically significant positive markers for all clusters, then filter for genes that are markers **only** in a single cluster.

```{r method_a_markers}
cat("Running FindAllMarkers (this may take a few minutes depending on dataset size)...\n")
# Run FindAllMarkers
all_markers <- FindAllMarkers(
  object = seurat_obj,
  only.pos = TRUE,
  min.pct = MIN_PCT,
  logfc.threshold = LOGFC_THRESH,
  test.use = "wilcox",
  group.by = "seurat_clusters_res0.75"
)

# Save all markers to CSV
write.csv(all_markers, file = file.path(OUTPUT_DIR, "all_markers.csv"), row.names = FALSE)

# Filter for significant markers
sig_markers <- all_markers %>%
  filter(p_val_adj < P_VAL_ADJ_THRESH)

# Identify genes that are markers for exactly one cluster
gene_marker_counts <- table(sig_markers$gene)
unique_genes <- names(gene_marker_counts[gene_marker_counts == 1])

# Subset marker table to only these uniquely assigned genes
unique_markers <- sig_markers %>%
  filter(gene %in% unique_genes)

# Save unique markers to CSV
write.csv(unique_markers, file = file.path(OUTPUT_DIR, "unique_markers_method_a.csv"), row.names = FALSE)

# Count unique markers per cluster
method_a_counts <- unique_markers %>%
  group_by(cluster) %>%
  summarize(num_unique_genes = n(), .groups = "drop")

# Make sure all clusters are represented, even if they have 0 unique markers
all_clusters <- levels(Idents(seurat_obj))
method_a_counts <- data.frame(cluster = all_clusters) %>%
  left_join(method_a_counts, by = "cluster") %>%
  mutate(num_unique_genes = replace_na(num_unique_genes, 0))

print("Method A (Marker-Based) counts per cluster:")
print(method_a_counts)
```

### Plot Method A Results

```{r method_a_plot, fig.width=10, fig.height=6}
p1 <- ggplot(method_a_counts, aes(x = cluster, y = num_unique_genes)) +
  geom_point(aes(size = num_unique_genes, color = num_unique_genes), alpha = 0.8) +
  geom_text(aes(label = num_unique_genes), vjust = -1.2, size = 3.5, fontface = "bold", show.legend = FALSE) +
  scale_color_gradientn(colors = c("#3182bd", "#8e31a3", "#de2d26")) +
  labs(
    title = "Method A: Number of Uniquely Expressed Marker Genes per Cluster",
    subtitle = paste0("Criteria: adj p-val < ", P_VAL_ADJ_THRESH, ", log2FC > ", LOGFC_THRESH, ", exclusive to one cluster"),
    x = "Seurat Cluster",
    y = "Number of Uniquely Expressed Genes",
    color = "Count",
    size = "Count"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey90", linetype = "dashed")
  ) +
  expand_limits(y = max(method_a_counts$num_unique_genes) * 1.15)

print(p1)

# Save plot to PDF
pdf(file.path(OUTPUT_DIR, "method_a_unique_genes_scatter.pdf"), width = 10, height = 6)
print(p1)
dev.off()
```

---

# Method B: Expression-Fraction-Based Uniquely Expressed Genes
Here, we compute the percentage of cells expressing each gene in each cluster. A gene is "uniquely expressed" in cluster $C$ if the fraction of cells expressing it in $C$ is $> P_{min}$ (e.g., 5%), regardless of expression in other clusters.

```{r method_b_expression_fraction}
cat("Calculating expression percentages per cluster...\n")
# Get expression data matrix
expr_mat <- GetAssayData(seurat_obj, assay = "RNA", slot = "data")
clusters <- Idents(seurat_obj)
unique_clusters <- levels(clusters)

# Calculate percentage of cells expressing each gene (expression > 0) per cluster
# Using sparse matrix rowSums is extremely fast
pct_mat <- matrix(0,
  nrow = nrow(expr_mat), ncol = length(unique_clusters),
  dimnames = list(rownames(expr_mat), unique_clusters)
)

for (clust in unique_clusters) {
  cells_in_clust <- which(clusters == clust)
  if (length(cells_in_clust) > 0) {
    # Count how many cells have non-zero expression for each gene
    non_zero_counts <- Matrix::rowSums(expr_mat[, cells_in_clust, drop = FALSE] > 0)
    pct_mat[, clust] <- non_zero_counts / length(cells_in_clust)
  }
}
pct_df <- as.data.frame(pct_mat)

# Save percentage matrix to CSV
write.csv(pct_df, file = file.path(OUTPUT_DIR, "percent_cells_expressing_per_cluster.csv"))

# Find genes matching target specificity criteria for each cluster
unique_genes_method_b <- list()
method_b_counts_list <- c()

for (clust in unique_clusters) {
  # Condition: target cluster expression > P_MIN (no restriction on other clusters)
  cond_target <- pct_df[[clust]] > P_MIN

  matching_genes <- rownames(pct_df)[cond_target]
  unique_genes_method_b[[clust]] <- matching_genes
  method_b_counts_list <- c(method_b_counts_list, length(matching_genes))
}

method_b_counts <- data.frame(
  cluster = unique_clusters,
  num_unique_genes = method_b_counts_list
)

# Write out the lists of uniquely expressed genes to a JSON or text files
sink(file.path(OUTPUT_DIR, "unique_genes_method_b.txt"))
for (clust in names(unique_genes_method_b)) {
  cat("========================================\n")
  cat("Cluster:", clust, "(", length(unique_genes_method_b[[clust]]), "genes)\n")
  cat("========================================\n")
  cat(paste(unique_genes_method_b[[clust]], collapse = ", "), "\n\n")
}
sink()

print("Method B (Expression-Fraction-Based) counts per cluster:")
print(method_b_counts)
```

### Plot Method B Results

```{r method_b_plot, fig.width=10, fig.height=6}
p2 <- ggplot(method_b_counts, aes(x = cluster, y = num_unique_genes)) +
  geom_point(aes(size = num_unique_genes, color = num_unique_genes), alpha = 0.8) +
  geom_text(aes(label = num_unique_genes), vjust = -1.2, size = 3.5, fontface = "bold", show.legend = FALSE) +
  scale_color_gradientn(colors = c("#3182bd", "#8e31a3", "#de2d26")) +
  labs(
    title = "Method B: Number of Expressed Genes per Cluster",
    subtitle = paste0("Criteria: > ", P_MIN * 100, "% cells in cluster (expression in other clusters not restricted)"),
    x = "Seurat Cluster",
    y = "Number of Expressed Genes",
    color = "Count",
    size = "Count"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey90", linetype = "dashed")
  ) +
  expand_limits(y = max(method_b_counts$num_unique_genes) * 1.15)

print(p2)

# Save plot to PDF
pdf(file.path(OUTPUT_DIR, "method_b_unique_genes_scatter.pdf"), width = 10, height = 6)
print(p2)
dev.off()
```

---

# Comparison of Methods
Let's see how the two definitions correlate.

```{r comparison, fig.width=8, fig.height=6}
comparison_df <- data.frame(
  cluster = method_a_counts$cluster,
  Method_A = method_a_counts$num_unique_genes,
  Method_B = method_b_counts$num_unique_genes
)

p_comp <- ggplot(comparison_df, aes(x = Method_A, y = Method_B)) +
  geom_point(size = 3, color = "darkblue", alpha = 0.7) +
  geom_text(aes(label = cluster), vjust = -0.8, hjust = 0.5, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Comparison of Gene Counts per Cluster",
    x = "Method A: Marker-based",
    y = "Method B: Expression-fraction-based"
  ) +
  theme_cowplot()

print(p_comp)

pdf(file.path(OUTPUT_DIR, "methods_comparison_scatter.pdf"), width = 8, height = 6)
print(p_comp)
dev.off()
```

---

# UMI Counts per Cluster
To examine the sequencing depth and UMI count distribution across the different clusters, we generate a violin plot showing the total UMI counts (`nCount_RNA`) per cell grouped by cluster.

```{r UMI_counts_plot, fig.width=10, fig.height=6}
# Create a dataframe for plotting
umi_df <- data.frame(
  cluster = Idents(seurat_obj),
  nCount_RNA = seurat_obj$nCount_RNA
)

# Sort the clusters numerically if possible, otherwise alphabetically
clust_vals <- unique(as.character(umi_df$cluster))
if (!any(is.na(suppressWarnings(as.numeric(clust_vals))))) {
  sorted_levels <- as.character(sort(as.numeric(clust_vals)))
  umi_df$cluster <- factor(umi_df$cluster, levels = sorted_levels)
} else {
  umi_df$cluster <- factor(umi_df$cluster, levels = sort(clust_vals))
}

# Plot violin of UMI counts per cluster
p_umi <- ggplot(umi_df, aes(x = cluster, y = nCount_RNA, fill = cluster)) +
  geom_violin(scale = "width", adjust = 1, trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.1, color = "black", alpha = 0.6, outlier.shape = NA) +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    title = "Total UMI Counts per Cluster",
    subtitle = "Distribution of total transcripts (nCount_RNA) per cell across clusters",
    x = "Seurat Cluster",
    y = "Total UMI Counts (nCount_RNA, log10 scale)"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    panel.grid.major.y = element_line(color = "grey90", linetype = "dashed")
  )

print(p_umi)

# Save plot to PDF
pdf(file.path(OUTPUT_DIR, "umi_counts_per_cluster_violin.pdf"), width = 10, height = 6)
print(p_umi)
dev.off()
```

---

# Mitochondrial Gene Expression per Cluster
To evaluate cell quality, stress, or metabolic activity across the clusters, we calculate and plot the percentage of mitochondrial gene expression per cell. Mitochondrial genes are identified as those starting with "MT" (or "mt").

```{r mitochondrial_expression_plot, fig.width=10, fig.height=6}
# Identify mitochondrial genes
mt_genes <- grep("^(MT|mt)", rownames(seurat_obj), value = TRUE)
cat("Found", length(mt_genes), "mitochondrial genes starting with MT/mt.\n")

if (length(mt_genes) > 0) {
  # Calculate percent.mt if not already present, or recalculate to ensure it matches the pattern
  seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, features = mt_genes)
  
  # Create a dataframe for plotting
  mt_df <- data.frame(
    cluster = Idents(seurat_obj),
    percent_mt = seurat_obj$percent.mt
  )
  
  # Sort the clusters numerically if possible, otherwise alphabetically
  clust_vals <- unique(as.character(mt_df$cluster))
  if (!any(is.na(suppressWarnings(as.numeric(clust_vals))))) {
    sorted_levels <- as.character(sort(as.numeric(clust_vals)))
    mt_df$cluster <- factor(mt_df$cluster, levels = sorted_levels)
  } else {
    mt_df$cluster <- factor(mt_df$cluster, levels = sort(clust_vals))
  }
  
  # Plot violin of percent mitochondrial expression per cluster
  p_mt <- ggplot(mt_df, aes(x = cluster, y = percent_mt, fill = cluster)) +
    geom_violin(scale = "width", adjust = 1, trim = TRUE, alpha = 0.7) +
    geom_boxplot(width = 0.1, color = "black", alpha = 0.6, outlier.shape = NA) +
    labs(
      title = "Mitochondrial Gene Expression per Cluster",
      subtitle = "Percentage of mitochondrial transcripts (percent.mt) per cell across clusters",
      x = "Seurat Cluster",
      y = "Mitochondrial Expression (%)"
    ) +
    theme_cowplot() +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "grey90", linetype = "dashed")
    )
  
  print(p_mt)
  
  # Save plot to PDF
  pdf(file.path(OUTPUT_DIR, "mitochondrial_expression_per_cluster_violin.pdf"), width = 10, height = 6)
  print(p_mt)
  dev.off()
} else {
  warning("No mitochondrial genes starting with MT/mt were found in the Seurat object. Plot was not generated.")
}
```

---

# Unique Genes vs Total UMI Counts per Cluster
To explore the relationship between cellular sequencing depth (total UMI counts) and the detection of cluster-specific uniquely expressed genes, we generate scatter plots for each cluster. For each cell in a cluster, we plot its total UMI count against the number of uniquely expressed genes (from Method A and Method B) detected (expression > 0) in that cell.

```{r unique_genes_vs_umi}
# Retrieve expression matrix
expr_mat <- GetAssayData(seurat_obj, assay = "RNA", slot = "data")
clusters <- Idents(seurat_obj)
unique_clusters <- levels(clusters)

# Sort the clusters numerically if possible, otherwise alphabetically
if (!any(is.na(suppressWarnings(as.numeric(unique_clusters))))) {
  sorted_clusters <- as.character(sort(as.numeric(unique_clusters)))
} else {
  sorted_clusters <- sort(unique_clusters)
}

plot_list <- list()

for (clust in sorted_clusters) {
  cells_in_clust <- which(clusters == clust)
  if (length(cells_in_clust) == 0) next

  cell_names_in_clust <- colnames(seurat_obj)[cells_in_clust]
  genes_b <- unique_genes_method_b[[clust]]
  umi_counts <- seurat_obj@meta.data[cell_names_in_clust, "nCount_RNA"]

  if (length(genes_b) > 0) {
    detected_b <- colSums(expr_mat[genes_b, cell_names_in_clust, drop = FALSE] > 0)
  } else {
    detected_b <- rep(0, length(cell_names_in_clust))
  }

  plot_df <- data.frame(umi_counts = umi_counts, detected_b = detected_b)

  plot_list[[clust]] <- ggplot(plot_df, aes(x = umi_counts, y = detected_b)) +
    geom_point(color = "#8e31a3", alpha = 0.6, size = 1) +
    geom_smooth(method = "lm", formula = y ~ x, color = "#de2d26", linetype = "dashed", se = FALSE) +
    labs(
      title = paste0("Cluster ", clust, " (", length(genes_b), " genes)"),
      x = "UMI Counts",
      y = "Unique Genes Detected"
    ) +
    theme_cowplot(font_size = 9) +
    theme(panel.grid.major = element_line(color = "grey95"))
}

# Save to PDF: 12 plots per page (3 columns x 4 rows)
PLOTS_PER_PAGE <- 12
pdf_path <- file.path(OUTPUT_DIR, "unique_genes_vs_umi_by_cluster.pdf")
valid_clusters <- sorted_clusters[!sapply(sorted_clusters, function(cl) is.null(plot_list[[cl]]))]
pages <- split(valid_clusters, ceiling(seq_along(valid_clusters) / PLOTS_PER_PAGE))

pdf(pdf_path, width = 8.5, height = 11)
for (page_clusters in pages) {
  page_plots <- plot_list[page_clusters]
  print(wrap_plots(page_plots, ncol = 3, nrow = 4))
}
dev.off()
cat("Saved scatter plots to:", pdf_path, "\n")

# Print pages in the Rmd report
for (page_clusters in pages) {
  print(wrap_plots(plot_list[page_clusters], ncol = 3, nrow = 4))
}
```

# Session Information
```{r session_info}
writeLines(capture.output(sessionInfo()), paste0(output, "sessionInfo.txt"))
```
