---
title: "CASSIA test on GAMM S2 cluster 1 subset (Docker)"
output: html_document
---

# Paths and setup for Docker

**Biomart (pig ↔ human gene names):**

- On the host (server):  
  `/w5home/bmoore/scRNAseq/GAMM/human_data/reh_cellrep_2020/Human_Pig_Biomart_Filtered_mod.txt`
- Inside Docker, mount that *directory* and use the path below:  
  `-v "/w5home/bmoore/scRNAseq/GAMM/human_data/reh_cellrep_2020:/biomart:ro"`  
  → file becomes `/biomart/Human_Pig_Biomart_Filtered_mod.txt`

**Example docker run:**

```bash
# Prefer --env-file over -e with a raw key (keeps the key out of shell history)
docker run -it --rm \
  --env-file /path/to/openai.env \
  -v "/w5home/rstewart/scRNAseq_downstream/shared_volume:/shared_volume" \
  -v "/w5home/rstewart/scRNAseq_downstream/src:/src" \
  -v "/w5home/bmoore/scRNAseq/GAMM/human_data/reh_cellrep_2020:/biomart:ro" \
  stewartlab/scrnaseq_downstream3:v1.2.3 /bin/bash

conda activate cassia_env
Rscript -e "rmarkdown::render('/src/cassiaRMSTestGammS2_clus1.Rmd', output_dir='/shared_volume/output_cassia_gammS2_clus1')"
```

**Important:** the cluster-1 subset starts as a single identity. This notebook
**reclusters** that subset (PCA → FindNeighbors → FindClusters) so markers/CASSIA
see multiple subclusters. Set `RECLUSTER_RESOLUTION` below as needed.

```{r paths}
# --- edit these if needed ---
OUTPUT_DIR <- "/shared_volume/output_cassia_gammS2_clus1"
SEURAT_RDS <- "/shared_volume/output_subset_20260714_183521/seurat.obj_subset.rds"
BIOMART_FILE <- "/biomart/Human_Pig_Biomart_Filtered_mod.txt"
# Cluster column used for markers + CASSIA (set automatically after recluster)
CLUSTER_COL <- "seurat_clusters"
CASSIA_OUT_NAME <- "gammS2_clus1_retina"
DIM.RED <- "umap"

# Reclustering (needed so FindAllMarkers / CASSIA have >1 cluster)
# Set FALSE to reuse seurat.obj_reclustered.rds already in OUTPUT_DIR
DO_RECLUSTER <- FALSE
RECLUSTER_RESOLUTION <- 0.5
RECLUSTER_DIMS <- 1:15
RECLUSTER_ALGORITHM <- 4  # 4 = Leiden in Seurat
# Set FALSE to reuse markers_merge_unique.csv already in OUTPUT_DIR
DO_FIND_MARKERS <- FALSE

# Set FALSE to reuse the most recent CASSIA_Pipeline_*/03_csv_files/*_FINAL_RESULTS.csv
# already in OUTPUT_DIR instead of re-running runCASSIA_pipeline() (avoids repeat
# OpenAI API calls/cost when only downstream steps need fixing)
DO_RUN_CASSIA <- FALSE

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
# Avoid setwd() here — it confuses knitr; use OUTPUT_DIR / file.path() instead
```

# load libraries
```{r libraries}
# Bind reticulate to the Docker conda env BEFORE library(CASSIA).
# CASSIA's .onLoad otherwise initializes a uv-managed Python via py_require(),
# and setup_cassia_env(method = "conda") then fails with "another version
# of Python has already been initialized".
cassia_python <- Sys.getenv(
  "RETICULATE_PYTHON",
  unset = "/opt/conda/envs/cassia_env/bin/python"
)
if (!file.exists(cassia_python)) {
  stop("CASSIA Python not found at: ", cassia_python)
}
Sys.setenv(RETICULATE_PYTHON = cassia_python)
options(
  CASSIA.env_name = "cassia_env",
  CASSIA.env_method = "conda",
  CASSIA.conda_env = "cassia_env"
)

# matplotlib/seaborn are baked into cassia_env in image tag v1.2.3+

library(reticulate)
library(CASSIA)
library(Seurat)
library(dplyr)
library(stringr)
library(patchwork)
library(ggplot2)

# Deadlock workaround: BatchProgressTracker's ascii-mode _render() calls
# print(), which reticulate routes back to R's main thread. The main thread
# is itself blocked inside as_completed() while workers run, so it never
# services that callback -- worker threads hang forever holding the
# tracker's lock waiting to report progress. Neutralize the renderer so
# progress reporting never touches reticulate's stdout bridge.
reticulate::py_run_string("
import CASSIA.core.progress_tracker as _pt
_pt.BatchProgressTracker._render = lambda self: None
")
```

```{r setup_env}
# Do not call setup_cassia_env(method = "conda") here — it would try to re-init
# Python after reticulate is already attached to cassia_env.
print(py_config())
py_cassia <- get("py_cassia", envir = asNamespace("CASSIA"))
if (is.null(py_cassia)) {
  stop("CASSIA Python module failed to load (py_cassia is NULL)")
}
message("CASSIA Python ready: ", py_config()$python)
```

```{r api}
# Pass key into the container with --env-file (preferred) or -e OPENAI_API_KEY=...
# persist = FALSE avoids writing the key into the container filesystem
api_key <- Sys.getenv("OPENAI_API_KEY")
if (!nzchar(api_key)) {
  stop("OPENAI_API_KEY is not set. Start the container with --env-file or -e OPENAI_API_KEY=...")
}
setLLMApiKey(api_key, provider = "openai", persist = FALSE)
```

# read in subsetted Seurat object
```{r readin}
reclustered_rds <- file.path(OUTPUT_DIR, "seurat.obj_reclustered.rds")
if (!DO_RECLUSTER && file.exists(reclustered_rds)) {
  message("Loading existing reclustered object: ", reclustered_rds)
  seurat.obj <- readRDS(reclustered_rds)
  CLUSTER_COL <- paste0("seurat_clusters_res", RECLUSTER_RESOLUTION)
} else {
  seurat.obj <- readRDS(SEURAT_RDS)
}
print(seurat.obj)
print(colnames(seurat.obj@meta.data))
if (CLUSTER_COL %in% colnames(seurat.obj@meta.data)) {
  print(table(seurat.obj[[CLUSTER_COL]], useNA = "ifany"))
}
```

# recluster subset (creates multiple subclusters for CASSIA)
```{r recluster}
if (DO_RECLUSTER) {
  message("Reclustering subset at resolution ", RECLUSTER_RESOLUTION)

  if (!"pca" %in% names(seurat.obj@reductions)) {
    message("No PCA found; running RunPCA on variable features")
    if (length(VariableFeatures(seurat.obj)) == 0) {
      seurat.obj <- FindVariableFeatures(seurat.obj)
    }
    seurat.obj <- RunPCA(seurat.obj, features = VariableFeatures(seurat.obj), verbose = FALSE)
  }

  seurat.obj <- FindNeighbors(
    seurat.obj,
    dims = RECLUSTER_DIMS,
    reduction = "pca",
    verbose = FALSE
  )

  CLUSTER_COL <- paste0("seurat_clusters_res", RECLUSTER_RESOLUTION)
  seurat.obj <- FindClusters(
    seurat.obj,
    resolution = RECLUSTER_RESOLUTION,
    algorithm = RECLUSTER_ALGORITHM,
    cluster.name = CLUSTER_COL,
    verbose = FALSE
  )

  # Optional: refresh UMAP on this subset's PCA for clearer plots
  if (!DIM.RED %in% names(seurat.obj@reductions) || DO_RECLUSTER) {
    seurat.obj <- RunUMAP(seurat.obj, dims = RECLUSTER_DIMS, reduction = "pca", verbose = FALSE)
    DIM.RED <- "umap"
  }

  message("New cluster column: ", CLUSTER_COL)
  print(table(seurat.obj[[CLUSTER_COL]]))

  pdf(file.path(OUTPUT_DIR, paste0("umap_recluster_", CLUSTER_COL, ".pdf")), width = 8, height = 6)
  print(DimPlot(seurat.obj, reduction = DIM.RED, group.by = CLUSTER_COL, label = TRUE, pt.size = 0.5))
  dev.off()

  saveRDS(seurat.obj, file = file.path(OUTPUT_DIR, "seurat.obj_reclustered.rds"))
  message("Saved reclustered object to ", file.path(OUTPUT_DIR, "seurat.obj_reclustered.rds"))
} else {
  message("Skipping recluster; using CLUSTER_COL = ", CLUSTER_COL)
}
```

# set idents to (re)cluster column
```{r set_idents}
if (!(CLUSTER_COL %in% colnames(seurat.obj@meta.data))) {
  stop(paste("CLUSTER_COL not found in metadata:", CLUSTER_COL))
}
Idents(object = seurat.obj) <- CLUSTER_COL
print(table(Idents(seurat.obj)))
n_clusters <- length(unique(as.character(Idents(seurat.obj))))
if (n_clusters < 2) {
  stop(paste(
    "Only", n_clusters, "cluster(s) after setup. FindAllMarkers/CASSIA need multiple clusters.",
    "Try a higher RECLUSTER_RESOLUTION (e.g. 0.8 or 1.0) or check DO_RECLUSTER."
  ))
}
```

# find markers for each cluster
```{r findmarkers}
markers_merged_file <- file.path(OUTPUT_DIR, "markers_merge_unique.csv")
if (!DO_FIND_MARKERS && file.exists(markers_merged_file)) {
  message("Skipping FindAllMarkers; will load ", markers_merged_file, " later")
  all.markers <- NULL
} else {
  all.markers <- FindAllMarkers(object = seurat.obj)
  head(all.markers)
  all.markers <- as.data.frame(all.markers)
  write.csv(all.markers, file = file.path(OUTPUT_DIR, "markers_celltypes_all.csv"), row.names = FALSE)
}
```

# map pig gene symbols to human via biomart table
```{r readin_geneid}
if (!DO_FIND_MARKERS && file.exists(file.path(OUTPUT_DIR, "markers_merge_unique.csv"))) {
  message("Skipping biomart remap; markers_merge_unique.csv already exists")
  gene_symbols_df <- NULL
} else {
  if (!file.exists(BIOMART_FILE)) {
    stop(paste(
      "Biomart file not found at:", BIOMART_FILE,
      "\nMount the host directory when starting Docker:",
      '\n  -v "/w5home/bmoore/scRNAseq/GAMM/human_data/reh_cellrep_2020:/biomart:ro"'
    ))
  }
  gene_symbols_df <- read.csv(file = BIOMART_FILE, header = TRUE, sep = "\t")
  print(head(gene_symbols_df))
  gene_symbols_df$pig.gene.name <- toupper(gene_symbols_df$pig.gene.name)
}
```

# merge with marker list
```{r merge_markers}
if (!DO_FIND_MARKERS && file.exists(file.path(OUTPUT_DIR, "markers_merge_unique.csv"))) {
  message("Skipping merge; using saved markers")
} else {
  markers_merge <- merge(all.markers, gene_symbols_df,
    by.x = "gene", by.y = "pig.gene.name",
    all = FALSE
  )
  print(head(markers_merge))
}
```

# rename columns and drop
```{r tidy_markers}
markers_merged_file <- file.path(OUTPUT_DIR, "markers_merge_unique.csv")
if (!DO_FIND_MARKERS && file.exists(markers_merged_file)) {
  markers_merge_unique <- read.csv(markers_merged_file, stringsAsFactors = FALSE)
  message("Loaded markers_merge_unique: ", nrow(markers_merge_unique), " rows")
} else {
  markers_merge$pig_gene <- markers_merge$gene
  markers_merge$gene <- markers_merge$human.gene.name
  print(colnames(markers_merge))
  markers_merge <- markers_merge[, 1:7]
  head(markers_merge)
  markers_merge_unique <- distinct(markers_merge)
  nrow(markers_merge_unique)
  nrow(markers_merge)
  write.csv(markers_merge_unique, file = markers_merged_file, row.names = FALSE)
}
```

# run CASSIA
```{r cassia}
if (DO_RUN_CASSIA) {
  runCASSIA_pipeline(
    output_file_name = CASSIA_OUT_NAME,
    output_dir = OUTPUT_DIR,
    tissue = "retina",
    species = "pig",
    marker = markers_merge_unique,
    max_workers = 4,
    # TEMPORARY: annotation merging deadlocks under reticulate (see _render
    # monkeypatch above -- merge_annotations_all has the same print()-from-
    # worker-thread problem, but isn't covered by that patch). Disabled for
    # now to unblock this test run; re-enable once merging is fixed/patched.
    do_merge_annotations = FALSE,
    annotation_model = "gpt-5.1",
    annotation_provider = "openai",
    score_model = "gpt-5.1",
    score_provider = "openai",
    annotationboost_model = "gpt-5.1",
    annotationboost_provider = "openai"
  )
} else {
  message("Skipping runCASSIA_pipeline; will reuse existing FINAL_RESULTS.csv from OUTPUT_DIR")
}
```

# add cassia annotations back into seurat object
```{r add_cassia}
# Find the most recent CASSIA_Pipeline_*/03_csv_files/*_FINAL_RESULTS.csv under
# OUTPUT_DIR rather than hardcoding a timestamped path.
final_results_candidates <- Sys.glob(file.path(
  OUTPUT_DIR, "CASSIA_Pipeline_*", "03_csv_files", "*_FINAL_RESULTS.csv"
))
cassia_results_path <- if (length(final_results_candidates) > 0) {
  final_results_candidates[order(file.info(final_results_candidates)$mtime, decreasing = TRUE)][1]
} else {
  file.path(OUTPUT_DIR, "NO_FINAL_RESULTS_FOUND.csv")
}
if (!file.exists(cassia_results_path)) {
  warning(paste(
    "No FINAL_RESULTS.csv found under", OUTPUT_DIR,
    "-- run runCASSIA_pipeline() first (set DO_RUN_CASSIA <- TRUE)."
  ))
} else {
  message("Using CASSIA results: ", cassia_results_path)
  seurat.obj <- add_cassia_to_seurat(
    seurat_obj = seurat.obj,
    cassia_results_path = cassia_results_path,
    cluster_col = CLUSTER_COL,
    cassia_cluster_col = "Cluster ID",
    prefix = "Cassia_clus1_"
  )
}
```

```{r save}
saveRDS(seurat.obj, file = file.path(OUTPUT_DIR, "GAMM_S2_clus1_cassia.rds"))
```

```{r plot_if_annotated}
cassia_cols <- grep("^Cassia_clus1_", colnames(seurat.obj@meta.data), value = TRUE)
if (length(cassia_cols) >= 1) {
  plots <- lapply(cassia_cols[seq_len(min(3, length(cassia_cols)))], function(col) {
    DimPlot(seurat.obj,
      reduction = DIM.RED, label = TRUE,
      pt.size = 0.5, group.by = col
    )
  })
  combined_plot <- wrap_plots(plots, ncol = length(plots))
  pdf(file = file.path(OUTPUT_DIR, "Cassia_annot_clus1_umap.pdf"), width = 8, height = 6)
  print(combined_plot)
  dev.off()
} else {
  message("No Cassia_* metadata columns yet; skip plotting until add_cassia_to_seurat succeeds.")
}
```

<!--
The original cassia.Rmd had a second zebrafish high-expr/DE + CASSIA block below.
That is intentionally omitted here; this notebook is for the GAMM S2 cluster-1 Docker test only.
-->
