---
title: "Cell chat interactions"
author: "Beth Moore"
date: "`r Sys.Date()`"
output: html_document
---
# load libraries
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
ptm <- Sys.time()
library(CellChat)
library(patchwork)
reticulate::use_python("/w5home/bmoore/.virtualenvs/cellchat/bin/python", required = T)
```
# create dir
```{r}
setwd("/w5home/bmoore/Pierre_sc_zebrafish/cellchat_0.75_clusters/")
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
data.dir <- paste0("./comparison_4wk_clusters_0.75_", timestamp)
dir.create(data.dir, mode = "0777", showWarnings = FALSE)
setwd(data.dir)
```
# load cell chat objects from previous cell chat run
```{r}
name <- "4wk_R_NR"
cellchat.NR <- readRDS("/w5home/bmoore/Pierre_sc_zebrafish/cellchat_0.75_clusters/seurat_mouse_annot_4WkNR_cellchat_20260625_103118/seurat_mouse_annot_4WkNR_cellchat_obj.rds")
cellchat.R <- readRDS("/w5home/bmoore/Pierre_sc_zebrafish/cellchat_0.75_clusters/seurat_mouse_annot_4WkR_cellchat_20260625_103117/seurat_mouse_annot_4WkR_cellchat_obj.rds")
object.list <- list(R = cellchat.R, NR = cellchat.NR)
table(cellchat.NR@idents)
table(cellchat.R@idents)
# define a positive dataset, i.e., the dataset with positive fold change against the other dataset
pos.dataset <- "NR"
neg.dataset <- "R"
fibroblast <- "13" # c(8)
not_fibroblast <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28") # c(1,2,3,4,5,6,7,9,10,11,12,13,14,15,16,17)
Tcell <- "9" # c(6)
not_Tcell <- c("1", "2", "3", "4", "5", "6", "7", "8", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28") # c(1,2,3,4,5,7,8,9,10,11,12,13,14,15,16,17)
```
# merge
```{r}
cellchat_merge <- mergeCellChat(object.list, add.names = names(object.list))
cellchat_merge
execution.time <- Sys.time() - ptm
print(as.numeric(execution.time, units = "secs"))
```
# if cell composition is different across samples
```{r}
# i think this is only if one cell type does not exist in the other sample

# lift up cell chat objects before merge - this makes their cell type composition similar
# group.new = levels(cellchat.NR@idents)
# cellchat.R <- liftCellChat(cellchat.R, group.new)
# object.list <- list(NR = cellchat.NR, R = cellchat.R)
# cellchat_merge <- mergeCellChat(object.list, add.names = names(object.list), cell.prefix = TRUE)
```
# save
```{r}
# Users can now export the merged CellChat object and the list of the two separate objects for later use
save(object.list, file = paste0("cellchat_object.list_", name, ".RData"))
save(cellchat_merge, file = paste0("cellchat_merged_", name, ".RData"))
```

# Identify altered interactions and cell populations
```{r}
# Compare the total number of interactions and interaction strength and among cell populations
ptm <- Sys.time()
gg1 <- compareInteractions(cellchat_merge, show.legend = F, group = c(2, 1))
gg2 <- compareInteractions(cellchat_merge, show.legend = F, group = c(2, 1), measure = "weight")
pdf(file = paste0(name, "_global_interactions_pos", pos.dataset, ".pdf"))
print(gg1 + gg2)
dev.off()

# circle plot
pdf(file = paste0(name, "_cell_diffinteractions_circ_pos", pos.dataset, ".pdf"))
par(mfrow = c(1, 2), xpd = TRUE)
netVisual_diffInteraction(cellchat_merge, weight.scale = T)
netVisual_diffInteraction(cellchat_merge, weight.scale = T, measure = "weight")
dev.off()
# heatmap
# Red represent increased signals in second dataset compared to the first
gg1 <- netVisual_heatmap(cellchat_merge, comparison = c(1, 2))
#> Do heatmap based on a merged object
gg2 <- netVisual_heatmap(cellchat_merge, measure = "weight", comparison = c(1, 2))
#> Do heatmap based on a merged object
pdf(file = paste0(name, "_cell_diffinteractions_heatmap_pos", pos.dataset, ".pdf"))
print(gg1 + gg2)
dev.off()
```

# Compare the major sources and targets
```{r}
#  Identify cell populations with significant changes in signals
num.link <- sapply(object.list, function(x) {
        rowSums(x@net$count) + colSums(x@net$count) - diag(x@net$count)
})
weight.MinMax <- c(min(num.link), max(num.link)) # control the dot size in the different datasets
gg <- list()
for (i in 1:length(object.list)) {
        gg[[i]] <- netAnalysis_signalingRole_scatter(object.list[[i]],
                title = names(object.list)[i], weight.MinMax = weight.MinMax
        )
}
pdf(file = paste0(name, "_sig_changes_each_dataset.pdf"))
patchwork::wrap_plots(plots = gg)
dev.off()
```

```{r}
# sig changes of specific cell populations
# actuall sig.exclude is for pathways not cell types
# Positive values indicate the increase in the second dataset (R) while negative values indicate the increase in the first dataset (NR)
gg1 <- netAnalysis_signalingChanges_scatter(cellchat_merge,
        idents.use =
                Tcell, comparison = c(2, 1)
) # "T_Cells_and_Neutrophils"
gg2 <- netAnalysis_signalingChanges_scatter(cellchat_merge,
        idents.use =
                fibroblast, comparison = c(2, 1)
) # "Fibroblasts"

pdf(file = paste0(name, "_sig_changes_Tcell-Fibroblast.pdf"), height = 6, width = 10)
patchwork::wrap_plots(plots = list(gg1, gg2))
dev.off()

gg1 <- netAnalysis_signalingChanges_scatter(cellchat_merge,
        idents.use =
                Tcell, # "T_Cells_and_Neutrophils"
        xlims = c(-0.04, 0.0005), ylims = c(-0.025, 0.0005),
        comparison = c(2, 1)
)
gg2 <- netAnalysis_signalingChanges_scatter(cellchat_merge,
        idents.use =
                fibroblast, # "Fibroblasts"
        xlims = c(-0.04, 0.0005), ylims = c(-0.025, 0.0005),
        comparison = c(2, 1)
)

pdf(file = paste0(name, "_sig_changes_Tcell-Fibroblast1.pdf"), height = 6, width = 10)
patchwork::wrap_plots(plots = list(gg1, gg2))
dev.off()
```
```{r}
gg1 <- netAnalysis_signalingChanges_scatter(cellchat_merge,
        idents.use =
                Tcell, # "T_Cells_and_Neutrophils"
        xlims = c(-0.0005, 0.00015), ylims = c(-0.001, 0.00005),
        comparison = c(2, 1)
)
gg2 <- netAnalysis_signalingChanges_scatter(cellchat_merge,
        idents.use =
                fibroblast, # "Fibroblasts"
        xlims = c(-0.00005, 0.00005), ylims = c(-0.00005, 0.00005),
        comparison = c(2, 1)
)

pdf(file = paste0(name, "_sig_changes_Tcell-Fibroblast2.pdf"), height = 6, width = 10)
patchwork::wrap_plots(plots = list(gg1, gg2))
dev.off()
```
```{r}
celltype.list <- unique(object.list[[neg.dataset]]@idents) # object.list$R@idents
celltype.listA <- as.vector(celltype.list[celltype.list != c("9", "13")])
celltype.listA
```
```{r}
# loop through to for each cell type
for (i in 1:length(celltype.listA)) {
        print(celltype.listA[i])
        # celltype.list.tmp <- as.vector(celltype.listA[celltype.listA != celltype.listA[i]])
        gg1 <- netAnalysis_signalingChanges_scatter(cellchat_merge,
                idents.use =
                        celltype.listA[i],
                comparison = c(2, 1)
        )
        pdf(file = paste0(name, "_sig_changes_", celltype.listA[i], ".pdf"), height = 6, width = 6)
        print(gg1)
        dev.off()
}
```

#  Identify altered signaling with distinct network architecture
```{r}
# functional similarity
cellchat_merge <- computeNetSimilarityPairwise(cellchat_merge,
        type = "functional"
)
#> Compute signaling network similarity for datasets 1 2
cellchat_merge <- netEmbedding(cellchat_merge, type = "functional")
#> Manifold learning of the signaling networks for datasets 1 2
cellchat_merge <- netClustering(cellchat_merge, type = "functional")
#> Classification learning of the signaling networks for datasets 1 2
# Visualization in 2D-space
pdf(file = paste0(name, "_sig_networks_functional_2D_plot.pdf"))
netVisual_embeddingPairwise(cellchat_merge,
        type = "functional",
        label.size = 3.5
        # comparison = c(2, 1)
)
dev.off()

pdf(file = paste0(name, "_sig_networks_functional_2D_plot_pathslabel.pdf"))
netVisual_embeddingPairwise(cellchat_merge,
        type = "functional",
        label.size = 3.5, top.label = 0.2
        # comparison = c(2, 1)
)
dev.off()

# zoom in
pdf(file = paste0(name, "_sig_networks_functional_2D_plot_zoomin.pdf"))
netVisual_embeddingPairwiseZoomIn(cellchat_merge, type = "functional", nCol = 2)
dev.off()
# # structural similarity
# cellchat_merge <- computeNetSimilarityPairwise(cellchat_merge,
#                 type = "structural")
# cellchat_merge <- netEmbedding(cellchat_merge, type = "structural")
# cellchat_merge <- netClustering(cellchat_merge, type = "structural")
# # Visualization in 2D-space
# pdf(file = paste0(name, "_sig_networks_structural_2D_plot.pdf"))
# netVisual_embeddingPairwise(cellchat_merge, type = "structural", label.size = 3.5)
# dev.off()
```
```{r}
# compute pathway distance in joint manifold
pdf(file = paste0(name, "_pathway_distance.pdf"))
rankSimilarity(cellchat_merge, type = "functional") # , comparison1 = c(2, 1)
dev.off()
```
# Identify altered signaling with distinct interaction strength
```{r}
# overall information flow of each signaling pathway or ligand-receptor pair
# Significant signaling pathways were ranked based on differences in the overall information flow
# overall info flow is defined by the sum of communication probability among all pairs of cell groups in
# the inferred network (i.e., the total weights)
gg1 <- rankNet(cellchat_merge,
        mode = "comparison", measure = "weight",
        sources.use = NULL, targets.use = NULL, stacked = T, do.stat = TRUE, comparison = c(2, 1),
        return.data = TRUE
)
gg2 <- rankNet(cellchat_merge,
        mode = "comparison", measure = "weight",
        sources.use = NULL, targets.use = NULL, stacked = F, do.stat = TRUE, comparison = c(2, 1),
        return.data = TRUE
)
df1 <- as.data.frame(gg1$signaling.contribution)
write.csv(df1, paste0(name, "_overall_info_flow_df.csv"), row.names = FALSE)
```
```{r}
gg1_obj <- gg1$gg.obj
gg2_obj <- gg2$gg.obj
pdf(file = paste0(name, "_overall_info_flow.pdf"))
print(gg1_obj + gg2_obj)
dev.off()
```
```{r}
# Compare outgoing (or incoming) signaling patterns associated with each cell population
library(ComplexHeatmap)
i <- 1
# combining all the identified signaling pathways from different datasets
# outgoing
pathway.union <- union(object.list[[i]]@netP$pathways, object.list[[i + 1]]@netP$pathways)
ht1 <- netAnalysis_signalingRole_heatmap(object.list[[i]],
        pattern = "outgoing", signaling =
                pathway.union, title = names(object.list)[i], width = 5, height = 6,
        font.size = 3, font.size.title = 7
)
ht2 <- netAnalysis_signalingRole_heatmap(object.list[[i + 1]],
        pattern = "outgoing", signaling =
                pathway.union, title = names(object.list)[i + 1], width = 5, height = 6,
        font.size = 3, font.size.title = 7
)
pdf(file = paste0(name, "_outgoing_sig_paths.pdf"))
draw(ht1 + ht2, ht_gap = unit(0.5, "cm"))
dev.off()
# incoming
ht3 <- netAnalysis_signalingRole_heatmap(object.list[[i]],
        pattern = "incoming", signaling =
                pathway.union, title = names(object.list)[i], width = 5, height = 6,
        color.heatmap = "GnBu", font.size = 3, font.size.title = 7
)
ht4 <- netAnalysis_signalingRole_heatmap(object.list[[i + 1]],
        pattern = "incoming", signaling =
                pathway.union, title = names(object.list)[i + 1], width = 5, height = 6,
        color.heatmap = "GnBu", font.size = 3, font.size.title = 7
)
pdf(file = paste0(name, "_incoming_sig_paths.pdf"))
draw(ht3 + ht4, ht_gap = unit(0.5, "cm"))
dev.off()
# all
ht5 <- netAnalysis_signalingRole_heatmap(object.list[[i]],
        pattern = "all", signaling =
                pathway.union, title = names(object.list)[i], width = 5, height = 6, color.heatmap = "OrRd",
        font.size = 3, font.size.title = 7
)
ht6 <- netAnalysis_signalingRole_heatmap(object.list[[i + 1]],
        pattern = "all", signaling =
                pathway.union, title = names(object.list)[i + 1], width = 5, height = 6, color.heatmap = "OrRd",
        font.size = 3, font.size.title = 7
)
pdf(file = paste0(name, "_all_sig_paths.pdf"))
draw(ht5 + ht6, ht_gap = unit(0.5, "cm"))
dev.off()
```
# Differential outgoing/incoming signaling heatmap (R vs NR)
CellChat's `netAnalysis_signalingRole_heatmap()` has no built-in "differential"
mode: it always plots a single object, and internally row-normalizes each
pathway to its own max (`sweep(mat, 1, apply(mat, 1, max), "/")`), so the two
side-by-side heatmaps above are only comparable in relative pattern, not raw
magnitude, between R and NR. `netVisual_heatmap()` *does* support a
differential view, but only for the overall interaction count/weight network
(`object@net$count`/`weight`), not the pathway-level centrality scores used
here.

This reimplements `netAnalysis_signalingRole_heatmap`'s internal
outgoing/incoming centrality matrix construction (see `centr <-
slot(object, slot.name)$centr`; `outdeg`/`indeg` per pathway), but for two
objects instead of one, aligns them to the same pathway and cluster sets
(zero-filling anything missing from one side), and takes the raw difference
`mat2 - mat1` -- the same "second minus first" convention already used by
`netVisual_heatmap()` in this file (see the "Red represent increased signals
in second dataset" comment above). Diverging color scale via `colorRamp3()`
(CellChat's re-export of `circlize::colorRamp2`): negative (comparison[1],
i.e. R, relatively higher) -> blue; positive (comparison[2], i.e. NR,
relatively higher) -> red.
```{r}
# Diverging color scale whose breakpoints are placed with an inverse
# power-law (v = frac^(1/power) * max) instead of a straight line from 0 to
# max. A plain 3-point linear ramp (0=white, max=full color) makes anything
# that isn't near the extreme max look almost white. Placing breakpoints
# more densely near 0 (power < 1) pushes small/faint differences toward full
# saturation much faster, so they're actually visible; power = 1 reproduces
# the old linear behavior. Positive and negative sides are scaled off their
# own max magnitude so one side isn't washed out by a larger swing on the
# other side.
build_diverging_ramp <- function(mat, color.diff, power = 0.5, n.steps = 10) {
        neg.max <- abs(min(min(mat), 0))
        pos.max <- max(max(mat), 0)
        frac <- (0:n.steps) / n.steps

        breaks <- numeric(0)
        colors <- character(0)

        if (neg.max > 0) {
                neg.breaks <- -rev(neg.max * frac^(1 / power))
                neg.colors <- rev(colorRampPalette(c("white", color.diff[1]))(n.steps + 1))
                breaks <- c(breaks, neg.breaks)
                colors <- c(colors, neg.colors)
        }
        if (pos.max > 0) {
                pos.breaks <- pos.max * frac^(1 / power)
                pos.colors <- colorRampPalette(c("white", color.diff[2]))(n.steps + 1)
                if (length(breaks) > 0) {
                        # drop the duplicate 0/white at the negative/positive junction
                        pos.breaks <- pos.breaks[-1]
                        pos.colors <- pos.colors[-1]
                }
                breaks <- c(breaks, pos.breaks)
                colors <- c(colors, pos.colors)
        }
        colorRamp3(breaks, colors)
}

netAnalysis_signalingRole_heatmap_diff <- function(object.list, comparison = c(1, 2),
                                                   pattern = c("outgoing", "incoming", "all"),
                                                   signaling = NULL, slot.name = "netP",
                                                   color.use = NULL,
                                                   color.diff = c("#2166ac", "#b2182b"),
                                                   contrast.power = 0.5,
                                                   title = NULL, width = 10, height = 8,
                                                   font.size = 8, font.size.title = 10,
                                                   cluster.rows = FALSE, cluster.cols = FALSE) {
        pattern <- match.arg(pattern)

        # build the raw pathway x cluster outgoing/incoming matrix for one
        # object, mirroring netAnalysis_signalingRole_heatmap's internals
        # (but skipping its per-row max normalization, since that would make
        # the two objects' values incomparable)
        get_mat <- function(object) {
                if (length(slot(object, slot.name)$centr) == 0) {
                        stop("Please run `netAnalysis_computeCentrality` on all objects first!")
                }
                centr <- slot(object, slot.name)$centr
                outgoing <- matrix(0, nrow = nlevels(object@idents), ncol = length(centr))
                incoming <- matrix(0, nrow = nlevels(object@idents), ncol = length(centr))
                dimnames(outgoing) <- list(levels(object@idents), names(centr))
                dimnames(incoming) <- dimnames(outgoing)
                for (j in seq_along(centr)) {
                        outgoing[, j] <- centr[[j]]$outdeg
                        incoming[, j] <- centr[[j]]$indeg
                }
                if (pattern == "outgoing") {
                        mat <- t(outgoing)
                } else if (pattern == "incoming") {
                        mat <- t(incoming)
                } else {
                        mat <- t(outgoing + incoming)
                }
                mat # rows = pathways, columns = clusters
        }

        obj1 <- object.list[[comparison[1]]]
        obj2 <- object.list[[comparison[2]]]
        mat1 <- get_mat(obj1)
        mat2 <- get_mat(obj2)

        # align to the same pathway (row) and cluster (column) sets, since
        # R/NR may not share identical cell-type composition or pathway sets
        if (is.null(signaling)) {
                signaling <- union(rownames(mat1), rownames(mat2))
        }
        clusters.union <- union(colnames(mat1), colnames(mat2))

        reindex <- function(mat) {
                out <- matrix(0,
                        nrow = length(signaling), ncol = length(clusters.union),
                        dimnames = list(signaling, clusters.union)
                )
                rows.keep <- rownames(mat)[rownames(mat) %in% signaling]
                cols.keep <- colnames(mat)[colnames(mat) %in% clusters.union]
                out[rows.keep, cols.keep] <- mat[rows.keep, cols.keep, drop = FALSE]
                out
        }
        mat1 <- reindex(mat1)
        mat2 <- reindex(mat2)

        # comparison[2] - comparison[1]: positive (red) = comparison[2] higher,
        # negative (blue) = comparison[1] higher
        mat.diff <- mat2 - mat1

        legend.name <- switch(pattern,
                outgoing = "Outgoing",
                incoming = "Incoming",
                all = "Overall"
        )
        if (is.null(title)) {
                title <- paste0(
                        "Differential ", tolower(legend.name), " signaling\n(",
                        names(object.list)[comparison[2]], " vs ", names(object.list)[comparison[1]], ")"
                )
        }

        if (is.null(color.use)) {
                color.use <- scPalette(ncol(mat.diff))
        }
        names(color.use) <- colnames(mat.diff)

        color.heatmap.use <- build_diverging_ramp(mat.diff, color.diff, power = contrast.power)

        df <- data.frame(group = colnames(mat.diff))
        rownames(df) <- colnames(mat.diff)

        col_annotation <- HeatmapAnnotation(
                df = df, col = list(group = color.use),
                which = "column", show_legend = FALSE, show_annotation_name = FALSE,
                simple_anno_size = grid::unit(0.2, "cm")
        )

        ha2 <- HeatmapAnnotation(Strength = anno_barplot(colSums(mat.diff),
                border = FALSE, gp = gpar(fill = color.use, col = color.use)
        ), show_annotation_name = FALSE)
        ha1 <- rowAnnotation(
                Strength = anno_barplot(rowSums(mat.diff), border = FALSE),
                show_annotation_name = FALSE
        )

        mat.plot <- mat.diff
        mat.plot[mat.plot == 0] <- NA

        df.heatmap <- as.data.frame(mat.plot)
        df1 <- as.data.frame(mat1)
        df2 <- as.data.frame(mat2)

        ht <- Heatmap(mat.plot,
                col = color.heatmap.use, na_col = "white",
                name = paste0(names(object.list)[comparison[2]], " - ", names(object.list)[comparison[1]]),
                bottom_annotation = col_annotation, top_annotation = ha2,
                right_annotation = ha1, cluster_rows = cluster.rows, cluster_columns = cluster.cols,
                row_names_side = "left", row_names_rot = 0,
                row_names_gp = gpar(fontsize = font.size), column_names_gp = gpar(fontsize = font.size),
                width = unit(width, "cm"), height = unit(height, "cm"),
                column_title = title, column_title_gp = gpar(fontsize = font.size.title),
                column_names_rot = 90,
                heatmap_legend_param = list(
                        title_gp = gpar(fontsize = 8, fontface = "plain"),
                        title_position = "leftcenter-rot", border = NA, legend_height = unit(20, "mm"),
                        labels_gp = gpar(fontsize = 8), grid_width = unit(2, "mm")
                )
        )
        return(list(ht = ht, df = df.heatmap, dfR = df1, dfNR = df2))
}
```
```{r}
# comparison = c(1, 2) matches object.list <- list(R = ..., NR = ...) above,
# i.e. names(object.list)[comparison] == c("R", "NR"), so diff = NR - R:
# blue = towards R, red = towards NR
res_diff_out <- netAnalysis_signalingRole_heatmap_diff(object.list,
        comparison = c(1, 2), pattern = "outgoing",
        signaling = pathway.union, width = 8, height = 10,
        font.size = 5, font.size.title = 8
)
ht_diff_out <- res_diff_out$ht
pdf(file = paste0(name, "_diff_outgoing_sig_paths.pdf"), width = 6, height = 9)
draw(ht_diff_out)
dev.off()
write.csv(res_diff_out$df, file = paste0(name, "_diff_outgoing_sig_paths.csv"), row.names = TRUE)
write.csv(res_diff_out$dfR, file = paste0("2wk_R_outgoing_sig_paths.csv"), row.names = TRUE)
write.csv(res_diff_out$dfNR, file = paste0("2wk_NR_outgoing_sig_paths.csv"), row.names = TRUE)

res_diff_in <- netAnalysis_signalingRole_heatmap_diff(object.list,
        comparison = c(1, 2), pattern = "incoming",
        signaling = pathway.union, width = 8, height = 10,
        font.size = 5, font.size.title = 8
)
ht_diff_in <- res_diff_in$ht
pdf(file = paste0(name, "_diff_incoming_sig_paths.pdf"), width = 6, height = 9)
draw(ht_diff_in)
dev.off()
write.csv(res_diff_in$df, file = paste0(name, "_diff_incoming_sig_paths.csv"), row.names = TRUE)
write.csv(res_diff_in$dfR, file = paste0("2wk_R_incoming_sig_paths.csv"), row.names = TRUE)
write.csv(res_diff_in$dfNR, file = paste0("2wk_NR_incoming_sig_paths.csv"), row.names = TRUE)
```
# Identify the up-gulated and down-regulated signaling ligand-receptor pairs
```{r}
# Id dysfunctional signaling by comparing the communication probabities
# for T cells
p1 <- netVisual_bubble(cellchat_merge,
        sources.use = Tcell,
        targets.use = not_Tcell, # sources.use = c(6), targets.use = c(1,2,3,4,5,7,8,9,10,11,12,13,14,15,16,17)
        comparison = c(2, 1), angle.x = 45
)
p2 <- netVisual_bubble(cellchat_merge,
        sources.use = not_Tcell,
        targets.use = Tcell, # sources.use = c(1,2,3,4,5,7,8,9,10,11,12,13,14,15,16,17)
        comparison = c(2, 1), angle.x = 45
)
pdf(
        file = paste0(name, "_sig_cc_L-Rpairs_celltypes_bubble_Tcells.pdf"),
        height = 11, width = 8.5
)
print(p1 + p2)
dev.off()
# fibroblasts
p3 <- netVisual_bubble(cellchat_merge,
        sources.use = fibroblast,
        targets.use = not_fibroblast, # sources.use = c(8), targets.use = c(1,2,3,4,5,6,7,9,10,11,12,13,14,15,16,17)
        comparison = c(2, 1), angle.x = 45
)
p4 <- netVisual_bubble(cellchat_merge,
        sources.use = not_fibroblast,
        targets.use = fibroblast,
        comparison = c(2, 1), angle.x = 45
)
pdf(
        file = paste0(name, "_sig_cc_L-Rpairs_celltypes_bubble_fibroblasts.pdf"),
        height = 11, width = 8.5
)
print(p3 + p4)
dev.off()
```

```{r}
# up-regulated (increased) and down-regulated (decreased) signaling ligand-receptor pairs between datasets
# based on communication probability
# Tcells
gg1 <- netVisual_bubble(cellchat_merge,
        sources.use = not_Tcell, targets.use = Tcell,
        comparison = c(2, 1), max.dataset = 2, title.name = "Increased signaling in R to Tcell", angle.x = 45, remove.isolate = T
)

gg2 <- netVisual_bubble(cellchat_merge,
        sources.use = not_Tcell, targets.use = Tcell,
        comparison = c(2, 1), max.dataset = 1, title.name = "Decreased signaling in R to Tcell", angle.x = 45, remove.isolate = T
)

gg3 <- netVisual_bubble(cellchat_merge,
        sources.use = Tcell, targets.use = not_Tcell,
        comparison = c(2, 1), max.dataset = 2, title.name = "Increased signaling in R from Tcell", angle.x = 45, remove.isolate = T
)

gg4 <- netVisual_bubble(cellchat_merge,
        sources.use = Tcell, targets.use = not_Tcell,
        comparison = c(2, 1), max.dataset = 1, title.name = "Decreased signaling in R from Tcell", angle.x = 45, remove.isolate = T
)

pdf(file = paste0(name, "_inc-dec_sig_cc_L-Rpairs_celltypes_bubble_Tcells.pdf"), height = 11, width = 8.5)
print(gg1 + gg2 + gg3 + gg4)
dev.off()

# Fibroblasts
gg5 <- netVisual_bubble(cellchat_merge,
        sources.use = not_fibroblast, targets.use = fibroblast,
        comparison = c(2, 1), max.dataset = 2, title.name = "Increased signaling in R to Fibroblast", angle.x = 45, remove.isolate = T
)

gg6 <- netVisual_bubble(cellchat_merge,
        sources.use = not_fibroblast, targets.use = fibroblast,
        comparison = c(2, 1), max.dataset = 1, title.name = "Decreased signaling in R to Fibroblast", angle.x = 45, remove.isolate = T
)

gg7 <- netVisual_bubble(cellchat_merge,
        sources.use = fibroblast, targets.use = not_fibroblast,
        comparison = c(2, 1), max.dataset = 2, title.name = "Increased signaling in R from Fibroblast", angle.x = 45, remove.isolate = T
)

gg8 <- netVisual_bubble(cellchat_merge,
        sources.use = fibroblast, targets.use = not_fibroblast,
        comparison = c(2, 1), max.dataset = 1, title.name = "Decreased signaling in R from Fibroblast", angle.x = 45, remove.isolate = T
)

pdf(file = paste0(name, "_inc-dec_sig_cc_L-Rpairs_celltypes_bubble_fibroblasts.pdf"), height = 11, width = 8.5)
print(gg5 + gg6 + gg7 + gg8)
dev.off()
```
```{r}
gg1_df <- as.data.frame(gg1$data)
gg2_df <- as.data.frame(gg2$data)
gg3_df <- as.data.frame(gg3$data)
gg4_df <- as.data.frame(gg4$data)
gg5_df <- as.data.frame(gg5$data)
gg6_df <- as.data.frame(gg6$data)
gg7_df <- as.data.frame(gg7$data)
gg8_df <- as.data.frame(gg8$data)
print("saving dataframes")
write.table(gg1_df,
        file = paste0(name, "_incR_sig_L-Rpairs_Tcells_target.txt"),
        sep = "\t", quote = FALSE
)
write.table(gg2_df,
        file = paste0(name, "_decR_sig_L-Rpairs_Tcells_target.txt"),
        sep = "\t", quote = FALSE
)
write.table(gg3_df,
        file = paste0(name, "_incR_sig_L-Rpairs_Tcells_source.txt"),
        sep = "\t", quote = FALSE
)
write.table(gg4_df,
        file = paste0(name, "_decR_sig_L-Rpairs_Tcells_source.txt"),
        sep = "\t", quote = FALSE
)
write.table(gg5_df,
        file = paste0(name, "_incR_sig_L-Rpairs_fibroblasts_target.txt"),
        sep = "\t", quote = FALSE
)
write.table(gg6_df,
        file = paste0(name, "_decR_sig_L-Rpairs_fibroblasts_target.txt"),
        sep = "\t", quote = FALSE
)
write.table(gg7_df,
        file = paste0(name, "_incR_sig_L-Rpairs_fibroblasts_source.txt"),
        sep = "\t", quote = FALSE
)
write.table(gg8_df,
        file = paste0(name, "_decR_sig_L-Rpairs_fibroblasts_source.txt"),
        sep = "\t", quote = FALSE
)
```
```{r}
# Identify dysfunctional signaling by using differential expression analysis
# define a char name used for storing the results of differential expression analysis
features.name <- paste0(pos.dataset, ".merged")
# perform differential expression analysis
cellchat_merge <- identifyOverExpressedGenes(cellchat_merge,
        group.dataset = "datasets",
        pos.dataset = pos.dataset, features.name = features.name, only.pos = FALSE,
        thresh.pc = 0.1, thresh.fc = 0.05, thresh.p = 0.05, group.DE.combined = FALSE
)
# map the results of differential expression analysis onto the inferred cell-cell communications
net <- netMappingDEG(cellchat_merge, features.name = features.name, variable.all = TRUE)
# extract the ligand-receptor pairs with upregulated ligands in LS
net.up <- subsetCommunication(cellchat_merge, net = net, datasets = pos.dataset, ligand.logFC = 0.05, receptor.logFC = NULL)
## extract the ligand-receptor pairs with upregulated ligands and upregulated receptors in NL, i.e.,downregulated in LS
net.down <- subsetCommunication(cellchat_merge, net = net, datasets = neg.dataset, ligand.logFC = -0.05, receptor.logFC = NULL)
# deconvolute signaling genes
gene.up <- extractGeneSubsetFromPair(net.up, cellchat_merge)
gene.down <- extractGeneSubsetFromPair(net.down, cellchat_merge)
```

```{r}
# Visualize the identified up-regulated and down-regulated signaling ligand-receptor pairs
# T-cell
pairLR.use.up <- net.up[, "interaction_name", drop = F]
gg1 <- netVisual_bubble(cellchat_merge,
        pairLR.use = pairLR.use.up, sources.use = Tcell,
        targets.use = not_Tcell, comparison = c(2, 1), angle.x = 90, remove.isolate = T,
        title.name = paste0("Up-regulated signaling in ", names(object.list)[2])
)

pairLR.use.down <- net.down[, "interaction_name", drop = F]
gg2 <- netVisual_bubble(cellchat_merge,
        pairLR.use = pairLR.use.down, sources.use = Tcell,
        targets.use = not_Tcell, comparison = c(2, 1), angle.x = 90, remove.isolate = T,
        title.name = paste0("Down-regulated signaling in ", names(object.list)[2])
)

pdf(file = paste0(name, "_up-down_reg_cc_L-Rpairs_celltypes_bubble_Tcells.pdf"))
print(gg1 + gg2)
dev.off()

gg3 <- netVisual_bubble(cellchat_merge,
        pairLR.use = pairLR.use.up, sources.use = fibroblast,
        targets.use = not_fibroblast, comparison = c(2, 1), angle.x = 90, remove.isolate = T,
        title.name = paste0("Up-regulated signaling in ", names(object.list)[2])
)
gg4 <- netVisual_bubble(cellchat_merge,
        pairLR.use = pairLR.use.down, sources.use = fibroblast,
        targets.use = not_fibroblast, comparison = c(2, 1), angle.x = 90, remove.isolate = T,
        title.name = paste0("Down-regulated signaling in ", names(object.list)[2])
)

pdf(file = paste0(name, "_up-down_reg_cc_L-Rpairs_celltypes_bubble_fibroblast.pdf"))
print(gg3 + gg4)
dev.off()
```

```{r}
# Wrapper: pre-filter a net data frame to the top N interactions by probability
# for the given source/target combination, then draw the chord diagram.
# This avoids circlize's gap.degree error when the full net has too many L-R pairs.
chord_top_net <- function(obj, sources, targets, net_df, n = 50, ...) {
        sub <- net_df[net_df$source %in% sources & net_df$target %in% targets, ]
        if (nrow(sub) == 0) {
                message("No interactions for this source/target combination — skipping")
                return(invisible(NULL))
        }
        net_top <- sub[order(sub$prob, decreasing = TRUE)[1:min(n, nrow(sub))], ]
        netVisual_chord_gene(obj,
                sources.use = sources, targets.use = targets,
                slot.name = "net", net = net_top, ...
        )
}

# Chord diagram
# T cells
pdf(file = paste0(name, "_up-down_reg_cc_L-Rpairs_celltypes_chord_Tcells.pdf"))
par(mfrow = c(2, 2), xpd = TRUE)
chord_top_net(object.list[[2]], Tcell, not_Tcell, net.up,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Up-regulated signaling in ", names(object.list)[2])
)
chord_top_net(object.list[[1]], Tcell, not_Tcell, net.down,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Down-regulated signaling in ", names(object.list)[2])
)
chord_top_net(object.list[[2]], not_Tcell, Tcell, net.up,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Up-regulated signaling in ", names(object.list)[2])
)
chord_top_net(object.list[[1]], not_Tcell, Tcell, net.down,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Down-regulated signaling in ", names(object.list)[2])
)
dev.off()

# fibroblasts
pdf(file = paste0(name, "_up-down_reg_cc_L-Rpairs_celltypes_chord_fibroblasts.pdf"))
par(mfrow = c(2, 2), xpd = TRUE)
chord_top_net(object.list[[2]], fibroblast, not_fibroblast, net.up,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Up-regulated signaling in ", names(object.list)[2])
)
chord_top_net(object.list[[1]], fibroblast, not_fibroblast, net.down,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Down-regulated signaling in ", names(object.list)[2])
)
chord_top_net(object.list[[2]], not_fibroblast, fibroblast, net.up,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Up-regulated signaling in ", names(object.list)[2])
)
chord_top_net(object.list[[1]], not_fibroblast, fibroblast, net.down,
        n = 50,
        lab.cex = 0.8, small.gap = 3.5,
        title.name = paste0("Down-regulated signaling in ", names(object.list)[2])
)
dev.off()
```
# Visually compare cell-cell communication 
```{r}
pdf(
        file = paste0(name, "_signal_cc_L-Rpairs_celltypes_chord_fromTcells.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
# compare all the interactions sending from T cells to  cells
for (i in 1:length(object.list)) {
        netVisual_chord_gene(object.list[[i]],
                sources.use = Tcell, targets.use = not_Tcell,
                lab.cex = 0.4, title.name = paste0("Signaling from Tcells - ", names(object.list)[i])
        )
}
dev.off()

pdf(
        file = paste0(name, "_signal_cc_L-Rpairs_celltypes_chord_toTcells.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
# compare all the interactions sending from T cells to  cells
for (i in 1:length(object.list)) {
        netVisual_chord_gene(object.list[[i]],
                sources.use = not_Tcell, targets.use = Tcell,
                lab.cex = 0.4, title.name = paste0("Signaling to Tcells- ", names(object.list)[i])
        )
}
dev.off()
```
```{r}
# Helper: return top-N L-R pair names ranked by summed interaction probability
# CellChat's pairLR.use validation has an operator precedence bug — it rejects
# a data frame with only interaction_name. Workaround: zero out non-top interactions
# in a copy of the prob array so pairLR.use is not needed at all.
filter_top_LR <- function(obj, sources, targets, n = 25) {
        prob <- obj@net$prob # [source, target, LR_pair]
        lr_strength <- apply(prob[sources, targets, , drop = FALSE], 3, sum)
        n_keep <- min(n, sum(lr_strength > 0))
        if (n_keep == 0) {
                return(obj)
        }
        keep_idx <- order(lr_strength, decreasing = TRUE)[1:n_keep]
        zero_idx <- setdiff(seq_len(dim(prob)[3]), keep_idx)
        obj@net$prob[, , zero_idx] <- 0
        obj
}

pdf(
        file = paste0(name, "_signal_cc_L-Rpairs_celltypes_chord_fromFibroblast.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
for (i in 1:length(object.list)) {
        obj_f <- filter_top_LR(object.list[[i]], fibroblast, not_fibroblast, n = 25)
        netVisual_chord_gene(obj_f,
                sources.use = fibroblast, targets.use = not_fibroblast,
                lab.cex = 0.2, small.gap = 0.5, big.gap = 5,
                title.name = paste0("Top 25 L-R from Fibroblast - ", names(object.list)[i])
        )
}
dev.off()

pdf(
        file = paste0(name, "_signal_cc_L-Rpairs_celltypes_chord_toFibroblast.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
for (i in 1:length(object.list)) {
        obj_f <- filter_top_LR(object.list[[i]], not_fibroblast, fibroblast, n = 25)
        netVisual_chord_gene(obj_f,
                sources.use = not_fibroblast, targets.use = fibroblast,
                lab.cex = 0.2, small.gap = 0.5, big.gap = 5,
                title.name = paste0("Top 25 L-R to Fibroblast - ", names(object.list)[i])
        )
}
dev.off()
```

```{r compare_paths}
# show all the significant signaling pathways from fibroblast
pdf(
        file = paste0(name, "_signal_cc_paths_celltypes_chord_fromFibroblast.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
for (i in 1:length(object.list)) {
        netVisual_chord_gene(object.list[[i]],
                sources.use = fibroblast, targets.use = not_fibroblast,
                slot.name = "netP", title.name = paste0(
                        "Signaling pathways sending from fibroblast - ",
                        names(object.list)[i]
                ), legend.pos.x = 10, lab.cex = 0.2, small.gap = 0.5, big.gap = 5
        )
}
dev.off()
# to fibroblast
pdf(
        file = paste0(name, "_signal_cc_paths_celltypes_chord_toFibroblast.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
for (i in 1:length(object.list)) {
        netVisual_chord_gene(object.list[[i]],
                sources.use = not_fibroblast, targets.use = fibroblast,
                slot.name = "netP", title.name = paste0(
                        "Signaling pathways sending to fibroblast - ",
                        names(object.list)[i]
                ), legend.pos.x = 10, lab.cex = 0.2, small.gap = 0.5, big.gap = 5
        )
}
dev.off()
# from Tcells
pdf(
        file = paste0(name, "_signal_cc_paths_celltypes_chord_fromTcells.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
# compare all the interactions sending from T cells
for (i in 1:length(object.list)) {
        netVisual_chord_gene(object.list[[i]],
                sources.use = Tcell, targets.use = not_Tcell,
                slot.name = "netP", title.name = paste0(
                        "Signaling pathways sending from Tcells - ",
                        names(object.list)[i]
                )
        )
}
dev.off()
# to Tcells
pdf(
        file = paste0(name, "_signal_cc_paths_celltypes_chord_toTcells.pdf"),
        height = 10, width = 10
)
par(mfrow = c(1, 2), xpd = TRUE)
# compare all the interactions sending from T cells
for (i in 1:length(object.list)) {
        netVisual_chord_gene(object.list[[i]],
                sources.use = not_Tcell, targets.use = Tcell,
                slot.name = "netP", title.name = paste0(
                        "Signaling pathways sending to Tcells - ",
                        names(object.list)[i]
                ),
                legend.pos.x = 10, lab.cex = 0.2, small.gap = 0.5, big.gap = 5
        )
}
dev.off()
```
```{r}
# detailed look at paths
pathways.show.all <- unique(c(object.list[[1]]@netP$pathways, object.list[[2]]@netP$pathways))

# pathways.show.all.fin <- unique(vctrs::vec_c(pathways.show.all, pathways.show.all2))
for (j in 1:length(pathways.show.all)) {
        print(pathways.show.all[j])
        tryCatch(
                {
                        weight.max <- getMaxWeight(object.list,
                                slot.name = c("netP"),
                                attribute = pathways.show.all[j]
                        ) # control the edge weights across different datasets
                        pdf(file = paste0(pathways.show.all[j], "_path_sig_networks_circle2.pdf"))
                        par(mfrow = c(1, 2), xpd = TRUE)
                        for (i in 1:length(object.list)) {
                                groupSize <- as.numeric(table(object.list[[i]]@idents))
                                netVisual_aggregate(object.list[[i]],
                                        signaling = pathways.show.all[j],
                                        layout = "circle", edge.weight.max = weight.max[1], edge.width.max = 10, vertex.weight = groupSize,
                                        signaling.name = paste(pathways.show.all[j], names(object.list)[i])
                                )
                        }
                        dev.off()
                },
                error = function(e) {
                        an.error.occurred <<- TRUE
                        message(paste("weight max error ", an.error.occurred))
                        message("Here's the original error message:")
                        message(conditionMessage(e))
                        tryCatch(
                                {
                                        groupSize <- as.numeric(table(object.list[[1]]@idents))
                                        pdf(file = paste0(pathways.show.all[j], "_path_sig_networks_circle2.pdf"))
                                        netVisual_aggregate(object.list[[1]],
                                                signaling = pathways.show.all[j],
                                                layout = "circle", edge.weight.max = weight.max[1], edge.width.max = 10, vertex.weight = groupSize,
                                                signaling.name = paste(pathways.show.all[j], names(object.list)[1])
                                        )
                                        dev.off()
                                },
                                error = function(e) {
                                        an.error.occurred <<- TRUE
                                        message(paste(an.error.occurred))
                                        message("Here's the original error message:")
                                        message(conditionMessage(e))
                                        groupSize <- as.numeric(table(object.list[[2]]@idents))
                                        pdf(file = paste0(pathways.show.all[j], "_path_sig_networks_circle2.pdf"))
                                        netVisual_aggregate(object.list[[2]],
                                                signaling = pathways.show.all[j],
                                                layout = "circle", edge.weight.max = weight.max[1], edge.width.max = 10, vertex.weight = groupSize,
                                                signaling.name = paste(pathways.show.all[j], names(object.list)[2])
                                        )
                                        dev.off()
                                }
                        )
                }
        )
}
```
# get aggregated weights for each cell type
```{r}
# sources are rows, targets are columns
df.netNR <- as.data.frame(object.list[[1]]@net$weight)
df.netR <- as.data.frame(object.list[[2]]@net$weight)
write.table(df.netNR, file = paste0(name, "_netaggregateNR_weights.txt"), quote = FALSE, row.names = FALSE)
write.table(df.netR, file = paste0(name, "_netaggregateR_weights.txt"), quote = FALSE, row.names = FALSE)
```
# pathways
```{r}
df.net.p <- subsetCommunication(cellchat.NR, slot.name = "netP")
df.net.pR <- subsetCommunication(cellchat.R, slot.name = "netP")
write.table(df.net.p, file = paste0(name, "_NR_cellchat_df_net_signal_paths.txt"), sep = "\t", quote = FALSE)
write.table(df.net.pR, file = paste0(name, "_R_cellchat_df_net_signal_paths.txt"), sep = "\t", quote = FALSE)
# combine
df.net.p.merged <- merge(df.net.p, df.net.pR, by = c("source", "target", "pathway_name"), all = TRUE, suffixes = c("_NR", "_R"))

# replace NA with 0
df.net.p.merged[is.na(df.net.p.merged)] <- 0
# get net probability (NR - R)
df.net.p.merged$net_prob <- df.net.p.merged$prob_NR - df.net.p.merged$prob_R
write.table(df.net.p.merged, file = paste0(name, "_cellchat_df_net_signal_paths.txt"), sep = "\t", quote = FALSE)
```
# Compare the signaling gene expression distribution between different datasets
```{r}
cellchat_merge@meta$datasets <- factor(cellchat_merge@meta$datasets, levels = c(neg.dataset, pos.dataset)) # set factor level
# pathways.show <- c("MPZ","ADGRL","NCAM","CDH","NRXN","EPHB","CADM","NEGR","NOTCH","NRG","APP","NCAM","PTPR","ADGRB","NRG","UNC5")
for (j in 1:length(pathways.show.all)) {
        p1 <- plotGeneExpression(cellchat_merge,
                signaling = pathways.show.all[j], split.by = "datasets",
                colors.ggplot = T, type = "violin"
        )
        pdf(file = paste0(pathways.show.all[j], "_geneExpr_violin.pdf"))
        print(p1)
        dev.off()
}
```
# save
```{r save}
save(object.list, file = paste0("cellchat_object.list_", name, ".RData"))
save(cellchat_merge, file = paste0("cellchat_merge_", name, ".RData"))
writeLines(capture.output(sessionInfo()), paste0(name, "_sessionInfo.txt"))
```

```{r load_data}
name <- "2wk_R_NR"
# define a positive dataset, i.e., the dataset with positive fold change against the other dataset
pos.dataset <- "NR"
neg.dataset <- "R"
fibroblast <- "13" # c(8)
not_fibroblast <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28") # c(1,2,3,4,5,6,7,9,10,11,12,13,14,15,16,17)
Tcell <- "9" # c(6)
not_Tcell <- c("1", "2", "3", "4", "5", "6", "7", "8", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28") # c(1,2,3,4,5,7,8,9,10,11,12,13,14,15,16,17)

load(paste0("/w5home/bmoore/Pierre_sc_zebrafish/cellchat_0.75_clusters/comparison_2wk_clusters_0.75_20260818_133055/cellchat_object.list_", name, ".RData"))
load(paste0("/w5home/bmoore/Pierre_sc_zebrafish/cellchat_0.75_clusters/comparison_2wk_clusters_0.75_20260818_133055/cellchat_merge_", name, ".RData"))
setwd("/w5home/bmoore/Pierre_sc_zebrafish/cellchat_0.75_clusters/comparison_2wk_clusters_0.75_20260818_133055/")
```
