---
title: "clustifyr"
author: "Beth Moore"
date: "2023-09-22"
output: html_document
---

# clustifyr analysis for automatic cell type annotation with marker list or reference
# To run clustifyr, you need clustered seurat object

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#### installations ####
# see scRNAseq_library/src/clustifyr_setup.sh or
# use create conda environment with environment_scRNAseq-clustifyr.yml
# 

### load libraries ###
library(reticulate)
use_condaenv("scRNAseq_best")
library(clustifyr)
library(ggplot2)
library(cowplot)
library(Seurat)
library(ComplexHeatmap)
### set variables ###
BETHS_GAMM_DATA_DIR <- "/w5home/bmoore/scRNAseq/GAMM/"
### set working directory and output ###
setwd(paste0(BETHS_GAMM_DATA_DIR, "GAMM_S2/"))
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
output <- paste0("output_clustifyr_", timestamp)
dir.create(output, showWarnings = FALSE)
output <- paste0(output, "/")
```
# load data set and marker list
```{r load_data}
# load data set
seurat.obj <- readRDS(file = "output_20230830_155530/seurat_obj_labeled.rds")
# load marker list
markers <- read.csv2("../../../scRNAseq_library/sc_pipeline/known_marker_lists/Gamm_lab_Consolidated_markerList.txt", sep="\t", header = TRUE)
```
# rename identities if necessary
```{r rename_idents}
# rename clusters if necessary
table(seurat.obj$orig.ident)
table(seurat.obj$seurat_clusters)
table(seurat.obj@active.ident)
seurat.obj <- RenameIdents(object = seurat.obj, `Muller Glia - Retinal Prog` = "Retinal Prog - Muller Glia",
                                `Cones - Pan PRs` = "Cones")
table(seurat.obj@active.ident)
seurat.obj <- AddMetaData(seurat.obj, metadata = seurat.obj@active.ident, col.name= "CellType")
table(seurat.obj$CellType)
```
# make marker dataframe and rename columns
```{r marker_df}
# subset
markers_df <- data.frame(markers$gene,markers$Cell.type)
# rename columns 
colnames(markers_df) <- c("gene","cluster")
```
# run clustifyr from marker list
```{r clustifyr}
# run clustify lists
# Available metrics for clustify_lists include: "pct", "jaccard"
list_res <- clustify_lists(
  input = seurat.obj,              # seurat object
  cluster_col = "seurat_clusters", # name of column in meta.data containing cell clusters
  marker = markers_df,             # list of known marker genes
  metric = "pct",                  # test to use for assigning cell types
  marker_inmatrix = FALSE,
  obj_out = FALSE
)
# View as heatmap, or plot_best_call
p1 <- plot_cor_heatmap(
        cor_mat = list_res,      # matrix of correlation coefficients from clustify_lists()
        cluster_rows = TRUE,     # cluster by row?
        cluster_columns = TRUE,  # cluster by column?
        legend_title = "pct"     # title of heatmap legend
)
pdf(paste0(output, "correlation_heatmap.pdf"), width = 8, height = 6)
print(p1)
dev.off()
# Call cell types
list_res2 <- cor_to_call(
  cor_mat = list_res,              # matrix correlation coefficients
  cluster_col = "seurat_clusters"  # name of column in meta.data containing cell clusters
)
```
# add clustifyr calls as metadata and visualize
```{r add_meta}
# Insert into metadata dataframe as "clustifyr_call" column
clust_call <- call_to_metadata(
  res = list_res2,                 # data.frame of called cell type for each cluster
  metadata = seurat.obj@meta.data, 
  cluster_col = "seurat_clusters", # name of column in meta.data containing cell clusters
  rename_prefix = "clustifyr_call"          # set a prefix for the new column
)
# add to seurat object metadata
seurat.obj <- AddMetaData(seurat.obj, metadata = clust_call)
# plot
pc <- DimPlot(seurat.obj, reduction = "umap", group.by = "clustifyr_call_type", label = TRUE,
              label.size = 3, repel = TRUE) + ggtitle("Clustifyr annotated labels") +
              guides(fill = guide_legend(label.theme = element_text(size = 8)))
pdf(paste0(output,"clustifyr_marker_annotation_umap.pdf"), width = 11, height = 6)
print(pc)
dev.off()
```
# save bject with clustifyr annotation
```{r save}
saveRDS(seurat.obj, file= paste0(output, "seurat.obj_clustifyr.rds"))
```
# using reference (expression matrix with annotation or seurat object) with clustifyr
```{r load_ref}
# load annotated ref data
ref.seurat<- readRDS(file = paste0(BETHS_GAMM_DATA_DIR, "human_data/reh_cellrep_2020/output_preprocess20230912_144047_cc/human_D205_umap.rds"))
table(ref.seurat$type)
# subset
ref.seurat<- subset(ref.seurat, subset = type != c('AC2'))
ref.seurat<- subset(ref.seurat, subset = type != c('miG'))
ref.seurat<- subset(ref.seurat, subset = type != c('T2'))
# get ref matrix
new_ref_matrix <- seurat_ref(
  seurat_object = ref.seurat,        # SeuratV3 object
  cluster_col = "type"    # name of column in meta.data containing cell identities
)
```
# run clustify
```{r run_clustify}
# run clustify (spearman corr is default)
res <- clustify(
  input = seurat.obj,       # a Seurat object
  ref_mat = new_ref_matrix,    # matrix of RNA-seq expression data for each cell type
  cluster_col = "seurat_clusters", # name of column in meta.data containing cell clusters
  obj_out = TRUE,      # output Seurat object with cell type inserted as "type" column
  rename_prefix= "clustify_pred",
  n_genes= 2000
)
table(res$clustify_pred_type)
```
# visualize
```{r visualize}
pc <- DimPlot(res, reduction = "umap", group.by = "clustify_pred_type", label = TRUE,
              label.size = 3, repel = TRUE) + ggtitle("Clustifyr predicted labels- spearmans") +
  guides(fill = guide_legend(label.theme = element_text(size = 8)))
pdf(paste0(output, "clustifyr_predicted_labels_umap.pdf"), width = 10, height = 6)
print(pc)
dev.off()
table(res$clustify_pred_r)
```
# save object
```{r save}
saveRDS(res, file= paste0(output, "seurat.obj_clustifyr.rds"))
```