---
title: "sc-comp"
author: "Beth Moore"
date: "2023-09-22"
output: html_document
---

# sccomp analysis for cell type composition comparisons
# To run sccomp you need processed and labeled data sets that are annotated the same way

# Load libraries, set wd, and source functions
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#### installations ####
# see scRNAseq_library/src/sccomp_env_setup.sh or
# use scRNAseq_library/environment_sccomp.yml to create an sccomp environment
## remember to set your Rterm in settings for remote to the conda path for sccomp!!!

### load libraries ###
library(reticulate)
use_condaenv("sccomp", required=TRUE)
# devtools::install_github("Oshlack/speckle")
library(sccomp)
library(Seurat)
library(tidyverse)
library(loo)
library(reticulate)
library(devtools)
library(speckle)
library(ggplot2)
library(limma)
### set variables ###
BETHS_GAMM_DATA_DIR <- "/w5home/bmoore/scRNAseq/GAMM/"
CROSS_SPECIES <- "yes" # are you comparing across species (human to pig- then yes) or within species (no)
COMP_TYPE <- "seuratmapping" # based on the annotation- did you use scpred, seuratmapping, or manual 
### set working directory and output ###
setwd(paste0(BETHS_GAMM_DATA_DIR, "human_data/cowan_cell_2020/"))
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
output <- paste0("output_cell_composition_", timestamp)
dir.create(output, showWarnings = FALSE)
output <- paste0(output, "/")
packageVersion("sccomp")
```
# Read in processed and labeled data
```{r read_in}
seurat.obj.1 <- readRDS(file = "output_seurat_mapping_20231218_161017/human_IMR90_cca_pred.rds") # reference
# annotation level for reference
annot_ref<- seurat.obj.1$cell_type3
seurat.obj.2 <- readRDS(file = "output_seurat_mapping_20231218_161017/gamms2.2_cca_pred.rds") # query
```
# check orig.ident and rename if necessary
```{r idents}
colnames(seurat.obj.2@meta.data)
table(seurat.obj.2$orig.ident)
Idents(object = seurat.obj.2) <- "orig.ident"
seurat.obj.2 <- RenameIdents(object = seurat.obj.2, `output_S2-2_mm_mt_GeneFull` = "pig", `gamm_S2` = "pig")
seurat.obj.2[["orig.ident"]] <- Idents(object = seurat.obj.2)
table(seurat.obj.2$orig.ident)
# for manual, may need tor rename celltype
#seurat.obj.2 <- RenameIdents(object = seurat.obj.2, `Muller Glia - Retinal Prog` = "Retinal Prog - Muller Glia")
#table(seurat.obj.2@active.ident)
saveRDS(seurat.obj.2, file =  "output_seurat_mapping_20231218_161017/gamms2.2_cca_pred.rds")
# rename celltype column for reference
seurat.obj.1$ref_type <- annot_ref
table(seurat.obj.1$ref_type)
```
# subset cell types if needed
```{r subset}
table(seurat.obj.2$predicted.id)
seurat.obj.2<- subset(seurat.obj.2, subset = orig.ident != 'gamm_S2')
table(seurat.obj.2@meta.data$orig.ident)
table(seurat.obj.2$predicted.id)
```
# check location of annotation
```{r check}
if (CROSS_SPECIES=="yes"){
    #human
    table(annot_ref)
    #pig
    if (COMP_TYPE=="scpred"){
        table(seurat.obj.2$scpred_prediction)
    } else if (COMP_TYPE=="seuratmapping") {
       table(seurat.obj.2$predicted.id)
    } else {
        print("COMP_TYPE should be scpred or seurat mapping for cross-species")
    }
} else {
    if (COMP_TYPE=="scpred"){
        table(seurat.obj.1$scpred_prediction)
        table(seurat.obj.2$scpred_prediction)
    } else if (COMP_TYPE=="seuratmapping") {
       table(seurat.obj.1$predicted.id)
       table(seurat.obj.2$predicted.id)
    } else if (COMP_TYPE=="manual"){
        table(seurat.obj.1@active.ident)
        table(seurat.obj.2@active.ident)
    } else {
        print("COMP_TYPE should be scpred or seurat mapping or manual")
    }
}
```
# add identities to metadata
```{r add_meta}
if (COMP_TYPE=="manual"){
    seurat.obj.1 <- AddMetaData(seurat.obj.1, metadata = seurat.obj.1@active.ident, col.name= "CellType")
    table(seurat.obj.1$CellType)
    seurat.obj.2 <- AddMetaData(seurat.obj.2, metadata = seurat.obj.2@active.ident, col.name= "CellType")
    table(seurat.obj.2$CellType)
}
```
# merge datasets
```{r merge}
seurat.combined <- merge(seurat.obj.1, y = seurat.obj.2, add.cell.ids = c("sample1", "sample2"), project = "combined")
head(colnames(seurat.combined))
table(seurat.combined$orig.ident)
unique(seurat.combined$orig.ident)
unique(sapply(X = strsplit(colnames(seurat.combined), split = "_"), FUN = "[", 1))
table(Idents(object = seurat.combined))
```
# set and rename identities in combined dataset
```{r set_idents}
# set identity
Idents(object = seurat.combined) <- "orig.ident"
 table(Idents(object = seurat.combined))
if (CROSS_SPECIES=="yes"){
    seurat.combined <- RenameIdents(object = seurat.combined, `human_F49B7` = "human")
} else {
    seurat.combined <- RenameIdents(object = seurat.combined, `output_S1-1_mm_mt_GeneFull` = "s1",
                                `output_S1-2_mm_mt_GeneFull` = "s1", `output_S2-1_mm_mt_GeneFull` = "s2",
                                `output_S2-2_mm_mt_GeneFull` = "s2")
}
# stash idents
seurat.combined[["idents"]] <- Idents(object = seurat.combined)
unique(seurat.combined$idents)
```
# remove NAs
```{r remove_NA}
# subset to remove NAs
seurat.combined <- subset(x= seurat.combined, idents != "NA")

if (CROSS_SPECIES=="yes") {
    if (COMP_TYPE=="scpred") {
    # scpred
    seurat.pig <- subset(x= seurat.combined, scpred_prediction != "NA")
    unique(seurat.pig$scpred_prediction)
    } else if (COMP_TYPE=="seuratmapping") {
      # seurat mapping
    seurat.pig <- subset(x= seurat.combined, predicted.id != "NA")
    unique(seurat.pig$predicted.id)
    } else {
        print("COMP_TYPE should be scpred or seurat mapping for cross-species")
    }
    # human
    seurat.human <- subset(x= seurat.combined, ref_type != "NA")
    unique(seurat.human$ref_type)
}
```
# plot cell type proportions
```{r plot_proportions}
if (CROSS_SPECIES == "yes") {
    if (COMP_TYPE=="scpred") {
    # scpred
    pig.plot <- plotCellTypeProps(x = seurat.pig, clusters = seurat.pig$scpred_prediction, sample = seurat.pig$orig.ident)
    } else if (COMP_TYPE=="seuratmapping") {
    # seurat mapping
    pig.plot <- plotCellTypeProps(x = seurat.pig, clusters = seurat.pig$predicted.id, sample = seurat.pig$orig.ident)
    } else {
        print("COMP_TYPE should be scpred or seurat mapping for cross-species")
    }
    # human
    hum.plot <- plotCellTypeProps(x = seurat.human, clusters = seurat.human$ref_type, sample = seurat.human$orig.ident)
    # save plot
    pdf(paste0(output, "celltype_proportions.pdf"), width = 8, height = 6)
    print(pig.plot + hum.plot)
    dev.off()
} else {
    prop.plot <- plotCellTypeProps(x = seurat.combined, clusters = seurat.combined$CellType, sample = seurat.combined$idents)
    pdf(paste0(output, "celltype_proportions.pdf"), width = 8, height = 6)
    print(prop.plot)
    dev.off()
}
```
# combine cell types
```{r combine_meta}
if (CROSS_SPECIES == "yes") {
    if (COMP_TYPE=="scpred") {
        New_idents <- c(seurat.human$ref_type,seurat.pig$scpred_prediction)
    } else if (COMP_TYPE=="seuratmapping") {
    New_idents <- c(seurat.human$ref_type,seurat.pig$predicted.id)
    } else {
        print("COMP_TYPE should be scpred or seurat mapping for cross-species")
    }
    seurat.combined@meta.data$"CellType" <- as.factor(New_idents)
}
table(seurat.combined$CellType)
```
# save combined object
```{r save}
saveRDS(seurat.combined, file = paste0(output,"seurat.combined.rds"))
```
# sccomp model composition
```{r model_comp}
# model composition
comp.tibble <- seurat.combined |>
  sccomp_glm( 
    formula_composition = ~ idents, 
    .sample =  orig.ident, 
    .cell_group = CellType, 
    bimodal_mean_variability_association = TRUE,
    cores = 1 
  )
# write table
comp.table <- as.data.frame(comp.tibble)
comp.table.1 <- comp.table[,1:7]
write.table(comp.table.1, file= paste0(output, "composition_result_table.txt"), quote = F, col.names = TRUE, row.names= F, sep= "\t")
# get counts from a specific result:
pr.table <- as.data.frame(comp.table["8","count_data"])
write.table(pr.table, file= paste0(output, "pr_composition_result_table.txt"), quote = F, col.names = TRUE, row.names= F, sep= "\t")
```
# compare models (Bayesian Anova)
```{r bayes_anova}
# need sccomp version 1.4 for this!
# Fit first model
model_with_factor_association = 
  seurat.combined |>
  sccomp_glm( 
    formula_composition = ~ idents, 
    .sample =  orig.ident, 
    .cell_group = CellType, 
    check_outliers = FALSE, 
    bimodal_mean_variability_association = TRUE,
    cores = 1, 
    enable_loo = TRUE
  )
# Fit second model
model_without_association = 
  seurat.combined |>
  sccomp_glm( 
    formula_composition = ~ 1, 
    .sample =  orig.ident, 
    .cell_group = CellType, 
    check_outliers = FALSE, 
    bimodal_mean_variability_association = TRUE,
    cores = 1 , 
    enable_loo = TRUE
  )
# Compare models
comparison <- loo_compare(
  model_with_factor_association |> attr("fit") |> loo(),
  model_without_association |> attr("fit") |> loo()
)
comp_df <- as.data.frame(comparison)
write.table(comp_df, file= paste0(output, "model_comparison_table.txt"), quote = F, col.names = TRUE, row.names= F, sep= "\t")
# is elpd_diff/se_diff > abs(5)?
c.res <- comp_df$elpd_diff[2]/comp_df$se_diff[2]
print(paste0("if ",c.res," is greater than abs(5), significant"))
```
# diffrential variability model
```{r diff_var_model}
res = 
  seurat.combined |>
  sccomp_glm( 
    formula_composition = ~ idents, 
    formula_variability = ~ idents,
    .sample = orig.ident,
    .cell_group = CellType,
    bimodal_mean_variability_association = TRUE,
    cores = 1 
  )
res.table <- as.data.frame(res)
res.table.1 <- res.table[,1:17]
write.table(res.table.1, file= paste0(output, "celltype_x_develop_composition_result_table.txt"), quote = F, col.names = TRUE, row.names= F, sep= "\t")
# plot results
plots = plot_summary(res)
# group proportion boxplot 
# blue box is posterior predictive check and colors are 
# associated with significant associations for composition and/or variability
pdf(paste0(output, "signif_associations_boxplot.pdf"), width = 8, height = 6)
print(plots$boxplot)
dev.off()
# credible interval plot 1
# error bars represent 95% credible interval
pdf(paste0(output, "credible_intervals_1d.pdf"), width = 8, height = 6)
print(plots$credible_intervals_1D)
dev.off()
# credible interval plot 2
pdf(paste0(output, "credible_intervals_2d.pdf"), width = 8, height = 6)
print(plots$credible_intervals_2D)
dev.off()
```
# save session info
```{r sessioninfo}
writeLines(capture.output(sessionInfo()), paste0(output,"sessionInfo.txt"))
```