# load packages
```{r load_packages}
# conda env scRNAseq_new
library(Seurat)
library(SeuratData)
library(patchwork)
library(scran)
library(BiocParallel)
library(cowplot)
library(reticulate)
library(purrr)
library(jsonlite)
library(rmarkdown)
library(ggplot2)
library(scater)
library(SingleCellExperiment)
library(tidyverse)
library(viridis)
```
# set dir and filenames
```{r dir_file}
GIT_DIR <- "/w5home/bmoore/scRNAseq_downstream/src/" # getwd()
DATA_DIR <- "/w5home/bmoore/Pierre_sc_zebrafish/"
filename <- "SCI6WK_PARSE_PG_SB_030123.rds"
setwd(DATA_DIR)
```
# read in
```{r readin_merged}
merged_seurat <- readRDS(file = paste0(DATA_DIR, filename))
print(merged_seurat)
colnames(merged_seurat@meta.data)
```
# create output
```{r}
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
output <- paste0(DATA_DIR, "output_expr_vis_", timestamp)
print(output)
dir.create(output, mode = "0777", showWarnings = FALSE)
output <- paste0(output, "/")
```
# get metadata for cell types
```{r metadata}
# clusterids <- read.csv(paste0(DATA_DIR, "celltype_annot_df.csv"), header = TRUE)
# print(clusterids)
# new_df_ordered <- clusterids[order(as.numeric(clusterids$Cluster)), ]
# new_cluster_ids <- new_df_ordered$Celltype
# print(new_cluster_ids)
# # make sure that idents are the clusters you want annotated
# merged_seurat <- SetIdent(merged_seurat, value = "seurat_clusters")
# # Rename the clusters based on the new IDs
# names(new_cluster_ids) <- levels(merged_seurat)
# merged_seurat <- RenameIdents(merged_seurat, new_cluster_ids)
# # put in CellType metadata
# merged_seurat$CellType <- Idents(merged_seurat)

# Separate Type into timepoint and condition
Type <- as.character(merged_seurat$Type)
# Split by "-"
split_res <- strsplit(Type, "-")
# Extract timepoint (first part)
merged_seurat$timepoint <- sapply(split_res, `[`, 1)
# Extract condition (second part, if exists)
merged_seurat$condition <- sapply(split_res, function(x) if (length(x) > 1) x[2] else NA)

# Verify
print(unique(merged_seurat$timepoint))
print(unique(merged_seurat$condition))

# Generate and plot the UMAP plot
# pdf(paste0(DATA_DIR, "labeled-clusters.pdf"), bg = "white")
# print(DimPlot(merged_seurat, reduction = "umap", label = TRUE,
#         pt.size = 0.5, group.by = "CellType"))
# dev.off()
```
# Pseudobulk analysis
```{r pseudobulk}
# pseudobulk cells by stimulation condition AND cell type
bulk_seurat <- AggregateExpression(merged_seurat,
    assays = "RNA",
    group.by = c("Type", "orig.ident", "seurat_clusters"),
    return.seurat = TRUE
) # , layer = "counts"
Cells(bulk_seurat)
# make new celltype column
bulk_seurat$celltype.comp <- paste(bulk_seurat$seurat_clusters, bulk_seurat$Type, sep = "_")
Idents(bulk_seurat) <- "celltype.comp"
unique(bulk_seurat$celltype.comp)
```
# modify expression matrix function
```{r}
modify_matrix <- function(df, type) {
    if (type == "bulk") {
        # Get the rownames (sample names)
        sample_names <- rownames(df)

        # Split by "_" and convert to a data.frame
        split_names <- do.call(rbind, strsplit(sample_names, "_"))

        # Assign to new columns in your dataframe
        df$timepoint <- split_names[, 1]
        df$rep <- split_names[, 2]
        df$celltype <- split_names[, 3]
    } else if (type == "merged") {
        sample_names <- as.character(df$ident)
        # Split by "_" and convert to a data.frame
        split_names <- do.call(rbind, strsplit(sample_names, "_"))
        df$timepoint <- split_names[, 2]
        df$celltype <- split_names[, 1]
    }
    print(head(df$timepoint))
    # split to get recovered/ not recovered
    R.NR.type <- do.call(rbind, strsplit(df$timepoint, "-"))[, 2]
    # print(head(R.NR.type))
    timepoint <- do.call(rbind, strsplit(df$timepoint, "-"))[, 1]
    timepoint <- ifelse(
        tolower(timepoint) == "sham",
        "0Wk",
        sub("^g(\\d+)Wk$", "\\1Wk", timepoint)
    )
    R.NR.type <- ifelse(
        R.NR.type %in% c("Sham", "g1Wk", "1Wk"), NA, R.NR.type
    )

    df$type <- R.NR.type
    df$timepoint <- timepoint

    # sort
    df <- df %>%
        arrange(timepoint)

    return(df)
}
```

```{r}
# Function to duplicate a given timepoint's NA row for each type
duplicate_timepoint_for_types <- function(df, tp, types) {
    rows <- df %>% filter(timepoint == tp & is.na(type))
    if (nrow(rows) == 0) {
        return(NULL)
    }
    expanded <- rows[rep(1, length(types)), ]
    expanded$type <- types
    expanded
}
```

# summarize function
```{r}
summarise_gene <- function(dfsub) {
    # get mean, min, max
    dfsub$rep <- as.numeric(as.character(dfsub$rep))

    # Summarize by timepoint and type
    df_summary <- dfsub %>%
        group_by(timepoint, type) %>%
        summarise(
            mean.expr = mean(norm.expr, na.rm = TRUE),
            min.rep = min(norm.expr, na.rm = TRUE),
            max.rep = max(norm.expr, na.rm = TRUE),
            .groups = "drop"
        )
    # Join summary back to dfsub if you want these columns in the original dfsub
    dfsub <- left_join(dfsub, df_summary, by = c("timepoint", "type"))

    # Identify all types to connect (e.g., NR, R)
    types_to_connect <- unique(df_summary$type[!is.na(df_summary$type)])

    # Duplicate for 0Wk and 1Wk
    zerowk_expanded <- duplicate_timepoint_for_types(df_summary, "0Wk", types_to_connect)
    onewk_expanded <- duplicate_timepoint_for_types(df_summary, "1Wk", types_to_connect)

    # Combine everything
    df_summary_connected <- df_summary %>%
        filter(!(timepoint %in% c("0Wk", "1Wk") & is.na(type))) %>%
        bind_rows(zerowk_expanded, onewk_expanded) %>%
        arrange(timepoint)

    return(df_summary_connected)
}
```
# plot function
```{r}
plot_function <- function(df_summary_connected, gene, Celltype, output) {
    p1 <- ggplot(data = df_summary_connected, aes(
        x = timepoint,
        y = mean.expr, group = type, color = type
    )) +
        geom_line(aes(color = type)) +
        geom_point() +
        geom_errorbar(aes(ymin = min.rep, ymax = max.rep), width = 0.1) +
        theme_minimal() +
        labs(
            title = paste0(gene, " ", Celltype), x = "Time point",
            y = "Mean normalized expr"
        )

    # make pdf
    pdf(
        file = paste0(
            output, gene, "_", Celltype, "_expr_lineplot",
            ".pdf"
        ),
        height = 4, width = 5
    )
    print(p1)
    dev.off()
}
```
# violin plot function
```{r}
vln_plot_function <- function(seurat.obj, gene, Celltype, output) {
    seurat.sub <- subset(seurat.obj, seurat_clusters == Celltype)

    # Handle NA conditions
    condition_vec <- as.character(seurat.sub$condition)
    condition_vec[is.na(condition_vec)] <- "both"
    seurat.sub$condition <- condition_vec

    # Fetch data for plotting
    plot_data <- FetchData(seurat.sub, vars = c(gene, "timepoint", "condition"))
    colnames(plot_data)[1] <- "expression"

    p1 <- ggplot(plot_data, aes(x = timepoint, y = expression, fill = condition)) +
        geom_violin(scale = "width", trim = TRUE) +
        geom_jitter(size = 0.1, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9)) +
        theme_minimal() +
        labs(
            title = paste0(gene, " ", Celltype), x = "Time point",
            y = "Normalized expr"
        )

    # make pdf
    pdf(
        file = paste0(
            output, gene, "_", Celltype, "_expr_violinplot",
            ".pdf"
        ),
        height = 4, width = 5
    )
    print(p1)
    dev.off()
}
```
# feature plot function
```{r}
feature_plot_function <- function(gene, timepoint, seurat.obj, output) {
    p1 <- FeaturePlot(seurat.obj,
        features = gene,
        pt.size = 0.1, reduction = "umap"
    ) &
        scale_color_viridis(option = "B")

    # make pdf
    pdf(
        file = paste0(
            output, gene, "_", timepoint, "_expr_featureplot",
            ".pdf"
        ),
        height = 4, width = 5
    )
    print(p1)
    dev.off()
}
```

# loop through cellchat files
```{r}
# need to loop through DE files here
de_dir <- file.path(DATA_DIR, "cellchat2")
filelist <- list.files(path = de_dir, pattern = "_target\\.txt$", full.names = TRUE)

for (f in filelist) {
    fname <- basename(f)
    fname_core <- sub("_target\\.txt$", "", fname)
    parts <- strsplit(fname_core, "_")[[1]]
    # celltype_part <- paste(parts[1:(length(parts) - 1)], collapse = "_")
    celltype_part <- parts[4]
    # comparison <- parts[length(parts)]
    timepoint_part <- parts[1]

    # Extract timepoint from celltype if present (e.g., "Macrophages6Wk")
    # This regex looks for a timepoint at the end of the celltype (e.g., "6Wk")
    timepoint_match <- regmatches(timepoint_part, regexpr("[0-9]+Wk", timepoint_part))
    timepoint <- ifelse(length(timepoint_match) > 0, timepoint_match, "1Wk-Sham")

    # Remove the timepoint from the celltype if it was found
    # Celltype <- ifelse(!is.na(timepoint),
    #   sub("[0-9]+Wk$", "", celltype_part),
    #   celltype_part
    # )

    # Read the file (adjust sep if needed)
    de_df <- read.table(f, header = FALSE, sep = "\t", stringsAsFactors = FALSE)

    print(paste("Celltype:", celltype_part, "Timepoint:", timepoint))

    if (is.null(de_df)) {
        print(paste0("No DE genes for ", celltype_part, " ", timepoint))
    } else {
        if (celltype_part == "Fibroblast") {
            Celltype <- "7"
        } else if (celltype_part == "Tcell") {
            Celltype <- "5"
        } else {
            Celltype <- celltype_part
        }
        # get genes
        gene.vector <- unique(unlist(strsplit(de_df$V1, "_")))
        # print(gene.vector)
        # read in orthologs
        orthologs <- read.table(file.path(DATA_DIR, "merged_orthologs_zebrafish-mouse.txt"), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
        # get zebrafish names
        orthologs <- orthologs %>% filter(Mouse.gene.name %in% gene.vector)
        zebrafish_names <- unique(orthologs$gene)

        # Filter for valid genes present in the object
        valid_genes <- intersect(zebrafish_names, rownames(bulk_seurat[["RNA"]]$counts))

        # get gene matrix
        # drop = FALSE ensures we keep dimensions even if only one gene is found
        gene.matrix <- bulk_seurat[["RNA"]]$counts[valid_genes, , drop = FALSE]

        # format data
        gene.matrix <- as.data.frame(t(gene.matrix))

        df <- modify_matrix(gene.matrix, "bulk")

        # loop through genes
        for (gene in valid_genes) {
            print(gene)
            # subset for gene and cell type
            dfsub <- df[, c(gene, "timepoint", "rep", "celltype", "type")]
            # print(head(dfsub))
            dfsub <- subset(x = dfsub, celltype %in% Celltype)
            colnames(dfsub) <- c("norm.expr", "timepoint", "rep", "celltype", "type")
            # summarise expression values
            df_summary <- summarise_gene(dfsub)
            # write summary
            write.csv(df_summary, file = paste0(
                output,
                "summary_",
                gene, "_", Celltype, "_", timepoint, ".csv"
            ))
            # plot summary
            plot_function(df_summary, gene, Celltype, timepoint, output)
        }
    }
}
```
# loop through genes and celltypes for lineplot of bulk expression
```{r}
# input gene vector and celltype
# gene.vector <- c("Cntn1", "Notch2", "Cntn2", "Nfasc", "Jag2", "App", "Cd74") #c("Hoxb5", "Etv4")
# Celltypes <- c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16")
gene.vector <- c("Hoxb5", "Etv4")
Celltypes <- c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16")
# read in orthologs
orthologs <- read.table(file.path(DATA_DIR, "merged_orthologs_zebrafish-mouse.txt"), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
# loop through celltypes
for (Celltype in Celltypes) {
    print(Celltype)
    # get zebrafish names
    orthologs <- orthologs %>% filter(Mouse.gene.name %in% gene.vector)
    zebrafish_names <- unique(orthologs$gene)

    # Filter for valid genes present in the object
    valid_genes <- intersect(zebrafish_names, rownames(bulk_seurat[["RNA"]]$counts))

    # get gene matrix
    # drop = FALSE ensures we keep dimensions even if only one gene is found
    gene.matrix <- bulk_seurat[["RNA"]]$counts[valid_genes, , drop = FALSE]

    # format data
    gene.matrix <- as.data.frame(t(gene.matrix))

    df <- modify_matrix(gene.matrix, "bulk")

    # loop through genes
    for (gene in valid_genes) {
        print(gene)
        # subset for gene and cell type
        dfsub <- df[, c(gene, "timepoint", "rep", "celltype", "type")]
        # print(head(dfsub))
        dfsub <- subset(x = dfsub, celltype %in% Celltype)
        colnames(dfsub) <- c("norm.expr", "timepoint", "rep", "celltype", "type")
        # summarise expression values
        df_summary <- summarise_gene(dfsub)
        # write summary
        write.csv(df_summary, file = paste0(
            output,
            "summary_",
            gene, "_", Celltype, ".csv"
        ))
        # plot summary
        plot_function(df_summary, gene, Celltype, output)
    }
}
```
# get genes and celltypes
```{r}
# input gene vector and celltype
gene.vector <- c("Hoxb5", "Etv4") # c("Hoxb5", "Etv4", "Cntn1", "Notch2", "Cntn2", "Nfasc", "Jag2", "App", "Cd74")
Celltypes <- c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16")
merged_seurat$celltype.comp <- paste(merged_seurat$seurat_clusters, merged_seurat$Type, sep = "_")
Idents(merged_seurat) <- "celltype.comp"
unique(merged_seurat$celltype.comp)
clusterassignments <- data.frame(ident = merged_seurat@active.ident)

# read in orthologs
orthologs <- read.table(file.path(DATA_DIR, "merged_orthologs_zebrafish-mouse.txt"),
    header = TRUE, sep = "\t", stringsAsFactors = FALSE
)
# get zebrafish names
orthologs <- orthologs %>% filter(Mouse.gene.name %in% gene.vector)
zebrafish_names <- unique(orthologs$gene)

# Filter for valid genes present in the object
valid_genes <- intersect(zebrafish_names, rownames(merged_seurat[["RNA"]]$counts))
```
# visualize expression over time with violin plots
```{r}
# loop through genes
for (gene in valid_genes) {
    print(gene)
    print("making violin plots")
    # loop through celltypes
    for (Celltype in Celltypes) {
        print(Celltype)
        # vioolin plot across time
        vln_plot_function(merged_seurat, gene, Celltype, output)
    }

    # feature plot
    print("making feature plots")
    for (type1 in unique(merged_seurat$Type)) {
        print(type1)
        # subset seurat object for timepoint
        seurat.sub <- subset(merged_seurat, Type == type1)
        feature_plot_function(gene, type1, seurat.sub, output)
    }
}
```

